---
title: 'OHI+ Palmyra 2020: 2_pal_spp_lists'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    toc_depth: 3 
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     in_header: '../../../src/templates/ohi_hdr.html' 
  pdf_document:
    toc: true
---

# Summary

Additional species lists were obtained for Palmyra fish and birds. These lists are added to the IUCN species list created in `1_iucn_spp_lists.Rmd`.   

# Data

The additional species lists were obtained from Alex Wegmann at The Nature Conservancy on July 8, 2020.   

# Setup 

```{r setup, message = F, warning = F}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(purrr)
library(rfishbase)
library(rredlist)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_goal   <- '~/github/pal-prep/prep/bd/spp/v2020'

# api keys for IUCN 
api_file <- file.path(dir_M, 'git-annex/globalprep/spp', 
                      'api_key_gc.csv')
api_key <- scan(api_file, what = 'character')
```

# Methods

## Palmyra Bird Lists

Some of the species on the bird list are believed extirpated on Palmyra, for these species the IUCN category will be updated to "EX" for extinct.   

### Get IUCN Info 

```{r pal-bird-list, eval=F}

birds <- read_csv(file.path(dir_goal, "raw/pal_bird_list.csv")) %>% 
  filter(sciname != is.na(sciname)) %>% 
  mutate(category = ifelse(status == "believed extirpated", "EX", category),
         category = ifelse(status == "extirpated", "EX", category)) %>%  # Re-categorize extirpated to EX
  dplyr::select(-status)

# Add class and iucn_sid using rredlist
# create list of scientific names of species
bird_scinames <- birds$sciname

iucn_info_birds <- data.frame() 

for(i in 1:length(bird_scinames)){
  
  sp <- bird_scinames[i]
  print(i)

  # Search IUCN reddlist for each species 
  result <- rredlist::rl_search(sp, key = api_key)$result
  
  # If species wasn't found, replace with NA
  if (length(result) == 0) {
    
    iucn_sid <- "NA"
    class    <- "NA"
    
  }
  
  if (length(result) != 0) {
   
    # Save relevant information if species was found
   iucn_sid <- result$taxonid
   class    <- result$class
   
  }
 
  # Create data frame
  df <- data.frame(iucn_sid = iucn_sid,
                   sciname = sp,
                   class = tolower(class))
  
  iucn_info_birds <- rbind(iucn_info_birds, df) %>%
    mutate(sciname = as.character(sciname))
}

# Add iucn info to the bird list
bird_list <- birds %>% 
  left_join(iucn_info_birds, by = "sciname")

```

### Add Missing Bird Info 

```{r add-missing-bird-info, eval=F}

# Manually add the missing IUCN species taxonomic ids
# Some weren't found because the scientific names in IUCN are different 
missing_ids <- data.frame(
  bird_scinames = c("Ardenna pacifica", "Anous ceruleus", "Onychoprion fuscatus", "Onychoprion lunatus"),
  common = c("Wedge-tailed shearwater", "Blue noddy", "Sooty tern", "Gray-backed tern"),
  missing_ids = c(132631353, 133493654, 132571340, 132569974)) %>% 
  mutate(common = as.character(common),
         bird_scinames = as.character(bird_scinames))

bird_list_filled <- bird_list %>% 
  left_join(missing_ids, by = "common") %>% 
  mutate(class = as.character(class),
         sciname = ifelse(iucn_sid == "NA", bird_scinames, sciname),
         iucn_sid = ifelse(iucn_sid == "NA", missing_ids, iucn_sid),
         class = ifelse(class == "na", "aves", class)) %>% 
  dplyr::select(iucn_sid, sciname, class, common, category)

# List of extirpated birds
ex_birds <- c("Puffinus lherminieri", "Ardenna pacifica", "Pterodroma alba",
              "Puffinus nativitatis", "Bulweria bulwerii", "Anous minutus",
              "Anous ceruleus", "Onychoprion lunatus")

# Save to int folder
write_csv(bird_list_filled, file.path(dir_goal, "int/bird_spp_list.csv"))
```


## Palmyra Fish List

### Get IUCN info 

```{r pal-fish-list, eval=F}

fish <- read_csv(file.path(dir_goal, "raw/PAL_KIN_fish_spp_list.csv")) %>% 
  dplyr::select(sciname = "TaxonName") %>% # Keep scientific names only
  
# Find the scientific names that don't have a species 
no_spp <- fish %>% 
  filter(str_detect(sciname, pattern = ". sp$")) %>% 
  .$sciname

# Write a "not contained in" function"
`%not_in%` <- purrr::negate(`%in%`)

# Filter fish list to remove the ones without species
fish_edit <- fish %>% 
  filter(sciname %not_in% no_spp) %>% 
  filter(sciname != "Acanthurus achillesXnigricans hybrid= rackliffei")

# Use rredlist to find common names, class and IUCN category
# create list of scientific names of species
fish_scinames <- fish_edit$sciname

iucn_info_fish <- data.frame() 

for(i in 1:length(fish_scinames)){
  
  sp <- scinames[i]
  print(i)

  # Search IUCN reddlist for each species 
  result <- rredlist::rl_search(sp, key = api_key)$result
  
  # If species wasn't found, replace with NA
  if (length(result) == 0) {
    
    iucn_sid <- "NA"
    class    <- "NA"
    comm     <- "NA"
    category <- "NA"  
    
  }
  
  if (length(result) != 0) {
   
    # Save relevant information if species was found
   iucn_sid <- result$taxonid
   class    <- result$class
   comm     <- result$main_common_name
   category <- result$category
   
  }
 
  # Create data frame
  df <- data.frame(iucn_sid = iucn_sid,
                   sciname = sp,
                   class = tolower(class),
                   common = comm,
                   category = category)
  
  iucn_info_fish <- rbind(iucn_info_fish, df) %>%
    mutate(common = as.character(common),
         sciname = as.character(sciname))
}

```

### Add Missing Fish Info

```{r find-missing-fish-info, eval=F}

# Find spp where the iucn_sid is NA
missing_fish_ids <- iucn_info_fish %>% 
  filter(iucn_sid == "NA") %>% # some scinames are not actually species 
  filter(sciname != "Data lost" & sciname != "no individuals seen" &
         sciname != "NO SCORE" & sciname != "Oxycheilinus sp." & 
         sciname != "Cheilinus sp.")

# Try and find common names in fishbase
fb_list <- species(missing_fish_ids$sciname) %>%
  dplyr::select(Species, FBname) %>% 
  filter(Species != is.na(Species),
         FBname != is.na(FBname)) %>% 
  rename(sciname = Species)

# These species are "Not Evaluated" by the IUCN 
missing_info <- missing_fish_ids %>% 
  left_join(fb_list, by = "sciname") %>% 
  mutate(category = "NE",
         class = "actinopterygii",
         class = as.character(class),
         category = as.character(category)) %>% # all species are not evaluated
  dplyr::select(-common) %>% 
  dplyr::select(iucn_sid, sciname, class, common = FBname, category)


# Add missing species info back to the fish list
fish_filled <- iucn_info_fish %>% 
  left_join(missing_info, by = "sciname") %>% 
  mutate(class.x = as.character(class.x),
         category.x = as.character(category.x),
         class = ifelse(class.x == "na", class.y, class.x),
         category = ifelse(category.x == "NA", category.y, category.x),
         common = ifelse(common.x == "NA", common.y, common.x)) %>% 
  filter(sciname != "Data lost" & sciname != "no individuals seen" &
         sciname != "NO SCORE" & sciname != "Oxycheilinus sp." & 
         sciname != "Cheilinus sp.") %>% # filter out the non species observations
  dplyr::select(iucn_sid = iucn_sid.x, sciname, class, common, category)


# Save to int folder
write_csv(fish_filled, file.path(dir_goal, "int/fish_spp_list.csv"))
```


## Combine Lists 

Cross reference the new bird and fish lists with the IUCN species list and add any species that weren't already on the list.  

```{r combine-lists, eval=F}

# Read in the three species lists
iucn_list <- read_csv(file.path(dir_goal, "int/iucn_spp_info.csv")) %>% 
  dplyr::select(-X1)
fish_list <- read_csv(file.path(dir_goal, "int/fish_spp_list.csv"))
bird_list <- read_csv(file.path(dir_goal, "int/bird_spp_list.csv"))

# Check if the species are included in the iucn spp list
setdiff(bird_list$sciname, iucn_list$sciname) # just one bird species isn't included
setdiff(fish_list$sciname, iucn_list$sciname) # 182 fish species weren't included 

# List of extirpated birds
ex_birds <- c("Puffinus lherminieri", "Ardenna pacifica", "Pterodroma alba",
              "Puffinus nativitatis", "Bulweria bulwerii", "Anous minutus",
              "Anous ceruleus", "Onychoprion lunatus")

# Combine lists
all_spp <- iucn_list %>% 
  full_join(bird_list, by = "sciname") %>% 
  mutate(iucn_sid.x = ifelse(is.na(iucn_sid.x), iucn_sid.y, iucn_sid.x),
         class.x = ifelse(is.na(class.x), class.y, class.x),
         common.x = ifelse(is.na(common.x), common.y, common.x),
         category.x = ifelse(sciname %in% ex_birds, "EX", category.x)) %>% 
  dplyr::select(-iucn_sid.y, -class.y, -category.y, -common.y) %>% 
  full_join(fish_list, by = "sciname") %>% 
  mutate(iucn_sid.x = ifelse(is.na(iucn_sid.x), iucn_sid, iucn_sid.x),
         class.x = ifelse(is.na(class.x), class, class.x),
         common.x = ifelse(is.na(common.x), common, common.x),
         category.x = ifelse(is.na(category.x), category, category.x)) %>% 
  dplyr::select(iucn_sid = iucn_sid.x, sciname, class = class.x, common = common.x, category = category.x)

# Save to int folder
write_csv(all_spp, file.path(dir_goal, "int/all_spp_list.csv"))
```

