---
title: 'Habitat Management'
subtitle: 'OHI+ Palmyra 2020 Assessment'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3 
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     in_header: '~/github/pal-prep/src/templates/pal_hdr.html' 
  pdf_document:
    toc: true
---

# Summary 

Regulatory resilience is meant to describes the institutional measures (e.g., rules, regulations, and laws) designed to address ecological pressures. This script calculates resilience from management of marine and terrestrial habitats on Palmyra Atoll.   

# Data 

Data to inform the preparation of these layers was obtained through verbal communications with Alex Wegmann of The Nature Conservancy on July 31, 2020.  

# Setup 

```{r setup, message=F, warning=F, results='hide'}

library(tidyverse)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_res   <- '~/github/pal-prep/prep/resilience/social/mgmt/v2020'
dir_lsp   <- '~/github/pal-prep/prep/sp/lsp/v2020'
dir_fp    <-  '~/github/pal-prep/prep/resilience/ecological/fp/v2020'
```


# Methods 

## Management 

A resilience layer is created for management of offshore and inland areas based on three criteria:  

 - Is there a management plan in place?  
 - Is there monitoring and assessment?   
 - Is the management plan effective?  
 
We will evaluate the terrestrial and marine areas separately. The marine management layer was created as part of the fishing pressure resilience score. The average of the marine and terrestrial scores will be used as a resilience for habitat destruction.  

A yes will be assigned a value of 1, in-development will be assigned a value of 0.5, a no will be assigned a value of 0. The final score will be the sum over 3. The status is assigned based on information from Alex Wegmann at The Nature Conservancy.  

Inland areas are managed by The Nature Conservancy. The Terrestrial Management Plan established in 2011 governs the management of the island and outlines the protocols for montioring and assessment. Management actions are updated based on results of the monitoring and assessment, this type of adaptive management tends to lead to more effective management.   
<br>   

| Category         | Response         | Score | 
|------------------|------------------|-------|
| Management Plan? | Yes              | 1     |
| Monitoring?      | Yes              | 1     |
| Effective?       | Yes              | 1     |

Offshore areas are managed by the U.S. Fish and Wildlife Service. A management plan is currently in development for the Palmyra Atoll National Wildlife Refuge. Currently, there isn't monitoring of the management plan as there is no management plan. However, NOAA does monitor coral reefs, oceanography parameters, and other key elements within the wildlife refuges on periodic scientific voyages. While this may not be the most robust form of management, the status of the area as fully protected and the remoteness have allowed for this to be fairly effective despite the absence of a formal management plan. For this assessment we will assign the effective category a score of 0.5.    
<br>   

| Category         | Response         | Score | 
|------------------|------------------|-------|
| Management Plan? | In Development   | 0.5   |
| Monitoring?      | Yes              | 1     |
| Effective?       | Yes              | 0.5   | 


```{r management, eval=F}

years <- 2014:2020

# Inland Management score
inland_mgmt <- data.frame(
  region_id = rep(1, length(years)), 
  year = years,
  dimension = rep("inland", length(years)),
  resilience_score = rep(3 / 3, length(years))
)

# Offshore Management score - read in from fishing pressure resilience 
offshore_mgmt <- read_csv(file.path(dir_fp, "output/fp_mgmt.csv")) 

# Average of management scores (for habitats)
hd_mgmt <- inland_mgmt %>% 
  rbind(offshore_mgmt) %>% 
  group_by(year) %>% 
  summarize(resilience_score = mean(resilience_score)) %>% 
  mutate(region_id = 1) %>% 
  dplyr::select(region_id, year, resilience_score)

# Save output layers
write_csv(hd_mgmt, file.path(dir_res, "output/hd_mgmt.csv"))
```

## Save to Toolbox

```{r save-to-toolbox, eval=F}
hd_mgmt <- read_csv(file.path(dir_res, "output/hd_mgmt.csv"))
write_csv(hd_mgmt, file.path(dir_scores, "layers/res_hd_mgmt.csv"))
```

