---
title: 'Habitat Destruction Regulations'
subtitle: 'OHI+ Palmyra 2020 Assessment'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3 
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     in_header: '~/github/pal-prep/src/templates/pal_hdr.html' 
  pdf_document:
    toc: true
---

# Summary 

Regulatory resilience is meant to describes the institutional measures (e.g., rules, regulations, and laws) designed to address ecological pressures. This script calculates resilience from regulations protecting marine and terrestrial habitats on Palmyra Atoll. Activities which would lead to habitat destruction, such as bottom trawling or deforestation, are prohibited in areas designated as protected.    

# Data 

Data to inform the preparation of these layers was obtained through verbal communications with Alex Wegmann of The Nature Conservancy on July 31, 2020.  

# Setup 

```{r setup, message=F, warning=F, results='hide'}

library(tidyverse)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_res   <- '~/github/pal-prep/prep/resilience/ecological/hd/v2020'
dir_lsp   <- '~/github/pal-prep/prep/sp/lsp/v2020'
dir_fp    <-  '~/github/pal-prep/prep/resilience/ecological/fp/v2020'
```


# Methods 

## Protected Areas

One regulatory mechanism to boost resilience to habitat destruction is designation of protected areas. The percent of protected areas is calculated as part of the Lasting Special Places sub-goal of the Sense of Place goal. For habitat destruction, we will average the scores for terrestrial and marine areas.    

```{r protected-areas, eval=F}

# Get offshore and inland LSP status scores 
offshore <- read_csv(file.path(dir_lsp, "output/offshore_status.csv")) %>% 
  mutate(dimension = "offshore")

inland   <- read_csv(file.path(dir_lsp, "output/inland_status.csv")) %>% 
  mutate(dimension = "inland")

# Use the average of the inland/offshore score for habitat destruction 
hd_pa <- offshore %>% 
  rbind(inland) %>% 
  group_by(year) %>% 
  summarize(status = mean(status)) %>% 
  filter(year >= 2014) %>% 
  mutate(region_id = 1) %>% 
  dplyr::select(region_id, year, resilience_score = status)

# Save files to output folder
write_csv(hd_pa, file.path(dir_res, "output/hd_reg.csv"))
```

## Save to Toolbox

```{r save-to-toolbox, eval=F}
hd_reg <- read_csv(file.path(dir_res, "output/hd_reg.csv"))
write_csv(hd_reg, file.path(dir_scores, "layers/res_hd_reg.csv"))
```

