---
title: 'OHI+ Palmyra 2020: Habitats Subgoal Coral Trend'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
     #in_header: '../../../workflow/templates/ohi_hdr.html' - need to update this
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

# Summary

This script estimates the trend in coral condition based on the OHI Global Assessment.   

# Setup

```{r setup, message = F, warning = F}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(ohicore)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_goal   <- '~/github/pal-prep/prep/bd/hab/v2020'
dir_global <- '~/github/ohiprep_v2020/globalprep'
```

# Data 

Coral condition is lacking historic data for all areas of the reef used to estimate current status (fore reef, back reef, and lagoon). Therefore, the trend of coral condition will be estimated using the OHI global data. The OHI global trend data is calculated as a linear trend in coral condition, calculated on a per country basis across all available data from 1975-2006. More information on the coral trend method can be found in the [OHI Global Methods](https://raw.githack.com/OHI-Science/ohi-global/published/documents/methods/Supplement.html#62_biodiversity) documentation, Section 6.2.1.      

# Methods

## Get Global Trend Data

```{r get-global-trend-data, eval=F}

# Global coral trend data
global_trend <- read_csv(file.path(dir_global, "hab_coral/v2012/data/habitat_trend_coral.csv"))

# Find region ids 
region_data() # Loads data frames needed to reference global region ids

pal_id <- rgns_all %>% 
  filter(rgn_name == "Palmyra Atoll") # Palmyra is region 150

# Check trend data for a Palmyra trend
pal_trend <- global_trend %>% 
  filter(rgn_id == 150) # No data - there is no trend estimate for Palmyra 

# Get UN georegions
UNgeorgn()

# Find Palmyra's georegions
pal_georgn <- UNgeorgn %>% 
  filter(rgn_label == "Palmyra Atoll") # Micronesia or Oceania 
```

## Compare Trend by Georegion

Global data resolution isn't the best and there is no estimate of coral trend for Palmyra Atoll. We will investigate the average trend in coral condition in both the Micronesia georegion and the Oceania georegion, as well as creating a region for just the U.S. Minor Outlying Islands.   

The [U.S. Minor Outlying Islands](https://www.iso.org/obp/ui#iso:code:3166:UM) in the Pacific Ocean consist of the following atolls and islands:  

 - Baker Island (Region 158)   
 - Howland Island (Region 158)   
 - Jarvis Island (Region 149)   
 - Johnston Atoll (Region 159)   
 - Kingman Reef (Not included)   
 - Midway Islands (Not included)   
 - Palmyra Atoll (Region 150)   
 - Wake Island (Region 12)   

```{r compare-trends, eval=F}

# Create a UM rgn_id vector 
um <- c(158,149,159,150,12)

# Add georegions
global_geo <- rgns_all %>% 
  left_join(UNgeorgn, by = "rgn_id") %>% 
  dplyr::select(rgn_id, rgn_name, r1_label, r2_label) %>% 
  mutate(um_island = ifelse(rgn_id %in% um, "US Minor Outlying Island", "no")) # Add UM designation

# Add coral trend data 
corals_georgn <- global_geo %>% 
  left_join(global_trend, by = "rgn_id") %>% 
  dplyr::select(-habitat)

# Find the different trends by Georegion
oceania <- corals_georgn %>% 
  filter(r1_label == "Oceania") %>% 
  group_by(r1_label) %>% 
  summarize(trend = mean(trend, na.rm=TRUE)) %>% 
  dplyr::select(geo_rgn = r1_label, trend)

micronesia <- corals_georgn %>% 
  filter(r2_label == "Micronesia") %>% 
  group_by(r2_label) %>% 
  summarize(trend = mean(trend, na.rm=TRUE)) %>% 
  dplyr::select(geo_rgn = r2_label, trend)

usmi <- corals_georgn %>% 
  filter(um_island == "US Minor Outlying Island") %>% 
  group_by(um_island) %>% 
  summarize(trend = mean(trend, na.rm=TRUE)) %>% 
  dplyr::select(geo_rgn = um_island, trend)

## Bind together to compare
compare_trends <- oceania %>% 
  rbind(micronesia) %>% 
  rbind(usmi)
```

The smallest geo-region is probably the most preferred. The US Minor Outlying Islands only has one island with a trend so instead we'll use the Micronesia georegion as the trend estimate.   

## Finalize Trend Data 

```{r finalize-trend-df, eval=F}

coral_trend <- micronesia %>% 
  mutate(region_id = 1,
         habitat = "coral",
         year = 2020) %>% 
  dplyr::select(region_id, habitat, year, trend)

# Save to output for use in cp trend
write_csv(coral_trend, file.path(dir_goal, "output/coral_trend.csv"))
```

## Save to Toolbox

```{r save-to-toolbox, eval=F}

# Save to toolbox
dir_scores <- '~/github/pal-scores/region'

write_csv(coral_trend, file.path(dir_scores, 'layers/hab_coral_trend.csv'))
```

