---
title: 'OHI+ Palmyra 2020: Habitats Subgoal Rainforest'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
     #in_header: '../../../workflow/templates/ohi_hdr.html' - need to update this
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

# Summary

This script uses data from a 2019 study quantifying coconut palm extent on the Pacific Islands to estimate the extent of the rainforest relative to what it could be if palm plantations on the island were removed.   

# Data 

 - Burnett, M.W. et al. (2019). "Quantifying coconut palm extent on Pacific islands using spectral textural analysis of very high resolution imagery". *International Journal of Remote Sensing* 40(19): 7329-7355   
  [Available here](https://www.tandfonline.com/doi/epub/10.1080/01431161.2019.1594440?needAccess=true). 
  
The data comes from Table A1 downloaded on July 8, 2020 from [here](https://www.tandfonline.com/doi/figure/10.1080/01431161.2019.1594440?scroll=top&needAccess=true).  


# Setup

```{r setup, message = F, warning = F}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(janitor)

# Source and set file paths
source(here('src/R/common.R'))

dir_goal   <- '~/github/pal-prep/prep/bd/hab/v2020'
dir_anx    <- file.path(dir_M, 'git-annex/foursite/palmyra')
```

# Methods

## Get Vegetation Data

Vegetation data is grouped into the following categories: cocos, native forest, scaelvola, and non-vegetated. Non-vegetated land area is excluded from the total area. Using the data we can find the total land area ($km^2$) and using the percent coverage, find the total area occupied by each vegetation category.  

```{r vegetation-data, eval=F}

veg_data <- read_csv(file.path(dir_anx, "_raw_data/Burnett_data/T0004-10.1080_01431161.2019.1594440.csv")) %>% 
  clean_names()

# Remove the (a) from the first column in total_area_m2
veg_data$total_area_m2 = gsub('a', "", veg_data$total_area_m2)

# Substitute , out of the total area column and convert to numeric
veg_data$total_area_m2 = as.numeric(gsub('[,]', "", veg_data$total_area_m2))

# Find the area of each vegetation type using the total area and the percent area
pal_veg <- veg_data %>% 
  mutate(total_area_km = total_area_m2 * 0.000001) %>% 
  dplyr::filter(islet == "Palmyra") # use the overall Palmyra atoll data

# Extract total area in km2
total_area <- pal_veg$total_area_km

# Extract the percent coverage by vegetation category 
veg_cover <- data.frame(category = c("cocos", "native_forest", "scaevola", "non_veg"), 
                        percent_cover = c(pal_veg$percent_cocos, pal_veg$percent_native_forest,
                                              pal_veg$percent_scaevola, pal_veg$percent_non_veg),
                        total_area_km = total_area) %>% 
  mutate(percent_cover = percent_cover / 100)

# Save to int folder
write_csv(veg_cover, file.path(dir_goal, "int/veg_cover.csv"))

```

## Calculate Vegetation Area

For this goal, we are concerned with calculating the area ($km^2$) of the rainforest and of coconut palm plantations

```{r calculate-percent-cover, eval=F}

veg_cover <- read_csv(file.path(dir_goal, "int/veg_cover.csv"))

# Regroup vegetation categories and find the total area 
veg_area <- veg_cover %>%
  mutate(category = ifelse(category == "scaevola", "forest", category),
         category = ifelse(category == "native_forest", "forest", category)) %>% # group the native forest and scaevola together as forest 
  mutate(cover_area = percent_cover * total_area_km) %>% 
  filter(category != "non_veg") %>% 
  dplyr::select(-total_area_km, - percent_cover) %>% 
  group_by(category) %>% 
  summarize(cover_area = sum(cover_area))

# Save to int folder
write_csv(veg_area, file.path(dir_goal, "int/veg_area.csv"))
```

## Get Historic Data

The historic percent coverage of palm and rainforest will be used as the reference point for the status score. Historic values are based on a 2001 satellite image evaluated in the U.S. Geological Survey 2011 Terrestrial Forest Management Plan for Palmyra Atoll. There is no downlaodable data, these numbers are pulled from the report to create the historic data frame.    

Citation: Hathaway, S.A., McEachern K., & Fisher, R.N. (2011). "Terrestrial Forest Management Plan for Palmyra Atoll". U.S. Geological Survey Open-File Report 2011-1007.  
  [Available here](https://permanent.fdlp.gov/gpo119875/ofr20111007.pdf)  
  
This report states that references to the size of the island vary, so a total land area value was used from 2000, which is the closest estimate of the area of the island when the 2001 satellite image was taken. The 2000 estimated land area was manually derived and cited in the 2019 study where the current status data was downloaded.  

The report gives the following vegetation categories and percent coverage:  

 - Cocos: 43%    
 - Scaevola: 29.5%   
 - Pisonia or native forest: 12%   
 - Broadleaf forest (combined with native forest): 6.2%   
 - Hau (combined with native forest): 0.5% 
 
The remaining 8.8% of the land area will be assumed to be non-vegetated.   
  
```{r historic-percent-cover, eval=F}

# Create a data frame from the values in the report 
historic_cover <- data.frame(category = c("cocos", "native_forest", "scaevola", "non_veg"),
                             percent_cover = c(0.43, 0.187, 0.295, 0.088),
                             total_area_km = 2.46) %>% 
  mutate(category = as.character(category))

# Regroup vegetation categories and find the total area 
hist_area <- historic_cover %>%
  mutate(category = ifelse(category == "scaevola", "forest", category),
         category = ifelse(category == "native_forest", "forest", category)) %>% # group the native forest and scaevola together as forest 
  mutate(cover_area = percent_cover * total_area_km) %>% 
  filter(category != "non_veg") %>% 
  dplyr::select(-total_area_km, - percent_cover) %>% 
  group_by(category) %>% 
  summarize(cover_area = sum(cover_area))

# Save to int folder
write_csv(hist_area, file.path(dir_goal, "int/historic_area.csv"))
```


## Calculate Status 

The rainforest habitat status is calculated based on the area of rainforest currently present compared to the area of rainforest that would exist once 95% of the  historic area of coconut palm has been removed. 

$status = area_{current}~(km^2)~/~area_{potential}~(km^2)$   

The current rainforest area ($area_{current}$) is based on the USGS mapping exercise from 2016 where forest is defined as the total area of the 9 vegetation communities listed above. 

The potential rainforest area ($area_{potential}$) is based on a 2001 satellite image showing approximately 43% of the islands are dominated by coconut plantations and 53% of the islands are dominated by other vegetation communities as discussed in the [USGS Terrestrial Management Plan](https://pubs.usgs.gov/of/2011/1007/pdf/ofr20111007.pdf) The potential rainforest area is calculated as the land area added once 95% of the historic palm plantations are removed plus the original 53% of land area that was dominated by rainforests.  

$area_{potential}~=~area_{added}~(km^2)~+~area_{original}~(km^2)$   

And where   

$area_{added}~=~0.95~*~(0.43~*~2.46~(km^2))$   
$area_{original}~=~0.53~*~2.46~(km^2)$   

```{r calculate-status, eval=F}

hist_area <- read_csv(file.path(dir_goal, "int/historic_area.csv"))
present_area <- read_csv(file.path(dir_goal, "int/veg_area.csv"))

# Find the area of land that would be added if 95% of historic coconut palms are removed
coco_area_hist <- hist_area$cover_area[hist_area$category == "cocos"]
area_added     <- 0.95 * coco_area_hist

# Total potential rainforest area is historic rainforest area + area added by removing 95% of coconut palms 
forest_area_hist      <- hist_area$cover_area[hist_area$category == "forest"]
potential_forest_area <- area_added + forest_area_hist 

# Define current forest area:
current_forest_area   <- present_area$cover_area[present_area$category == "forest"]

# Find status based on current forest area relative to potential forest area 
forest_status <- current_forest_area / potential_forest_area

# Create a final status data frame for export
rainforest_status <- data.frame(region_id = 1,
                                year = 2020,
                                status = forest_status)
```


## Save to Toolbox 


