---
title: 'OHI+ Palmyra 2020: Habitats Subgoal Rainforest Trend'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    toc_depth: 3 
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     in_header: '../../../src/templates/ohi_hdr.html' 
  pdf_document:
    toc: true
---

# Summary 

This script calculates the trend in rainforest area from 2005 to 2019 relative to the 2001 potential rainforest area.   

# Data 

Trend comes from three sources of satellite images and published papers:  

2001 Reference Point: The 2001 reference point comes from a satellite image of the island that is discussed in the 2011 Terrestrial Forest Management Plan.     
 - Hathaway, S.A., McEachern K., & Fisher, R.N. (2011). "Terrestrial Forest Management Plan for Palmyra Atoll". U.S. Geological Survey Open-File Report 2011-1007.  
  [Available here](https://permanent.fdlp.gov/gpo119875/ofr20111007.pdf)  
  
2005: The 2005 data was obtained from Alex Wegmann at The Nature Conservancy on July 31, 2020. Shapefiles were compiled as part of Alex Wegmann's PhD dissertation.  
 
2010: The 2010 data is estimated from a study on invasive rats that quantified the percent of land area on each islet occupied by *Cocos nucifera* (Table 1).    
 - Lafferty, K.D. et al. (2010). "Stomach nematodes (*Mastophorus muris*) in rats (*Rattus rattus*) are associated with coconut (*Cocos nucifera*) habitat at Palmyra Atoll. *Journal of Parasitology* 96(1): 16-20.   
  [Available here](https://bioone.org/journals/Journal-of-Parasitology/volume-96/issue-1/GE-2180.1/Stomach-Nematodes-Mastophorus-muris-in-Rats-Rattus-rattus-Are-Associated/10.1645/GE-2180.1.full)   
  
2016: The 2016 data is estimated using a U.S. Geological Survey vegetation community mapping project for Palmyra Atoll. Vegetation maps were completed in October 2016 and published in June 2019. Data was downloaded on July 2, 2020 from the USGS data portal, [here](https://www.sciencebase.gov/catalog/item/5c66fc5be4b0fe48cb3aca49).   
 - Struckhoff, M.A., 2019, Woody species crown and vegetation community data, Palmyra Atoll, 2016: U.S. Geological Survey data release, https://doi.org/10.5066/P9GMPVIO.  
 
 2019: The 2019 data set comes from Table A1 in the Burnett et al. paper   
  - Burnett, M.W. et al. (2019). "Quantifying coconut palm extent on Pacific islands using spectral textural analysis of very high resolution imagery". *International Journal of Remote Sensing* 40(19): 7329-7355   
  [Available here](https://www.tandfonline.com/doi/epub/10.1080/01431161.2019.1594440?needAccess=true). 

# Setup 

```{r setup, message = F, warning = F}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(sf)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_goal   <- '~/github/pal-prep/prep/bd/hab/v2020'
```


# Methods 

## 2001 Data

The historic percent coverage of palm and rainforest will be used as the reference point for the status score. Historic values are based on a 2001 satellite image evaluated in the U.S. Geological Survey 2011 Terrestrial Forest Management Plan for Palmyra Atoll. There is no downloadable data, these numbers are pulled from the report to create the data.    

This report states that references to the size of the island vary, so a total land area value was used from 2000, which is the closest estimate of the area of the island when the 2001 satellite image was taken. The 2000 estimated land area was manually derived and cited in the 2019 study where the current status data was downloaded.  

The report gives the following vegetation categories and percent coverage:  

 - Cocos: 43%    
 - Scaevola: 29.5%   
 - Pisonia or native forest: 12%   
 - Broadleaf forest (combined with native forest): 6.2%   
 - Hau (combined with native forest): 0.5% 
 
The remaining 8.8% of the land area will be assumed to be non-vegetated.  

Historic data was created into a data frame in the `rainforest_data_prep.Rmd` and saved as `historic_area.csv`.        

```{r 2001-area, eval=F}

raw_2001 <- read_csv(file.path(dir_goal, "int/historic_area.csv"))

# All data should be region_id, year, and area_forest
area_2001 <- raw_2001 %>% 
  filter(category == "forest") %>% 
  rename("area_forest" = "cover_area") %>% 
  mutate(region_id = 1, 
         year = 2001) %>% 
  dplyr::select(region_id, year, area_forest)

# Save to int folder
write_csv(area_2001, file.path(dir_goal, "int/trend_area_2001.csv"))
```

## 2005 Data 

Data for 2005 was obtained from Alex Wegmann at The Nature Conservancy. Shapefiles are included for cocos, pisonia, terminalia, scaevola, pandanus, hibiscus, grassland, and non-vegetated areas. All shapefiles are combined to produce a single data file for the 2005 data. Nonvegetated areas and grasslands will be excluded from the analysis as the focus of this goal is on forest communities.   

### Get 2005 Vegetation Data 

```{r get-2005-data, eval=F}

# All 2005 data
cocos_2005  <- st_read(file.path(dir_anx, '_raw_data/Wegmann_data/Palmyra_Forest_Project_ArcView_Shapefiles'),
                   layer = "cocos_forest")
hib_2005    <- st_read(file.path(dir_anx, '_raw_data/Wegmann_data/Palmyra_Forest_Project_ArcView_Shapefiles'),
                   layer = "HIBISC~1")
pand_2005   <- st_read(file.path(dir_anx, '_raw_data/Wegmann_data/Palmyra_Forest_Project_ArcView_Shapefiles'),
                   layer = "PANDAN~1")
pis_2005    <- st_read(file.path(dir_anx, '_raw_data/Wegmann_data/Palmyra_Forest_Project_ArcView_Shapefiles'),
                   layer = "pisonia_forest")
scaev_2005  <- st_read(file.path(dir_anx, '_raw_data/Wegmann_data/Palmyra_Forest_Project_ArcView_Shapefiles'),
                   layer = "SCAEVO~1")
term_2005   <- st_read(file.path(dir_anx, '_raw_data/Wegmann_data/Palmyra_Forest_Project_ArcView_Shapefiles'),
                   layer = "terminalia_forest")
grass_2005  <- st_read(file.path(dir_anx, '_raw_data/Wegmann_data/Palmyra_Forest_Project_ArcView_Shapefiles'),
                   layer = "GRASSL~1")
nonveg_2005 <- st_read(file.path(dir_anx, '_raw_data/Wegmann_data/Palmyra_Forest_Project_ArcView_Shapefiles'),
                   layer = "PHYMAT~1")

# Bind all the files together 
veg_2005 <- cocos_2005 %>% 
  rbind(hib_2005) %>% 
  rbind(pand_2005) %>% 
  rbind(pis_2005) %>% 
  rbind(scaev_2005) %>% 
  rbind(term_2005) %>% 
  rbind(grass_2005) %>% 
  rbind(nonveg_2005) %>% 
  dplyr::select(vegetation = VEGETATION, area = AREA)

# Save combined data to raw data folder
st_write(veg_2005, file.path(dir_goal, "raw/veg_communities_2005.shp"), driver = 'ESRI Shapefile', append=FALSE)
```

### Find Area 

```{r find-2005-area, eval=F}

veg_2005 <- st_read(file.path(dir_goal, "raw"), layer = "veg_communities_2005")

area_2005 <- veg_2005 %>% 
  filter(vegetation != "Grassland or Bare Ground" & vegetation != "Phymatosorus or Bare Ground") %>% 
  mutate(community = ifelse(vegetation == "Cocos", "cocos", "forest")) %>% 
  group_by(community) %>% 
  summarize(area = sum(area)) %>% 
  mutate(area_km = area * 0.000001,
         region_id = 1,
         year = 2005) %>% 
  filter(community == "forest") %>% 
  dplyr::select(region_id, year, area_forest = area_km)

# Convert to data frame and remove geometry
area_2005 <- as.data.frame(area_2005) %>% 
  dplyr::select(-geometry)

# Save to int folder 
write_csv(area_2005, file.path(dir_goal, "int/trend_area_2005.csv"))
```

### Grassland Area

Grasslands will be excluded from the analysis but we need to find grassland area to use as an estimate of grassland area in 2010.  

```{r grassland-area-2005, eval=F}
veg_2005 <- st_read(file.path(dir_goal, "raw"), layer = "veg_communities_2005")

# Find grassland area
grass_area_2005 <- veg_2005 %>% 
  filter(vegetation == "Grassland or Bare Ground") %>% 
  group_by(vegetation) %>% 
  summarize(area = sum(area))

# Find total area of 2005 study 
total_area_2005 <- sum(veg_2005$area)

# Find percent grassland 
pct_grass_2005 <- grass_area_2005$area / total_area_2005 # 8.77% 
```

### Bare Ground Area

Bare ground will be excluded from the analysis but we need to find the bare ground area to use as an estimate of bare ground in 2010.  

```{r bareground-area-2005, eval=F}
veg_2005 <- st_read(file.path(dir_goal, "raw"), layer = "veg_communities_2005")

# Find bare ground area
bare_2005 <- veg_2005 %>% 
  filter(vegetation == "Phymatosorus or Bare Ground") %>% 
  group_by(vegetation) %>% 
  summarize(area = sum(area))

# Find total area of 2005 study 
total_area_2005 <- sum(veg_2005$area)

# Find percent grassland 
pct_bare_2005 <- bare_2005$area / total_area_2005 # 3.66% 
```


## 2010 Data 

Data for 2010 is recreated from Table 1 in the Lafferty et al paper. The table gives each islet, the islet area (ha), and the percent cocos. Since no specific percent area of non-vegetated land is included in this study, we will use the average percent between the 2001, 2005, and 2016 data. The 2001 estimate for non-vegetated land is 8.8%, the 2005 estimate is 3.66% and the 2016 estimate is 4.69%. For the 2010 data, we'll assume 5.72% of the land area is non-vegetated. Grasslands have also been excluded for this analysis and since no specific percent of grasslands is given in the study we will use the average area of grasslands between 2005 and 2016. The 2005 estimate for grasslands is 8.77% and the 2016 estimate is 6.40%. For the 2010 data, we'll assume 7.59% of land area is grasslands.  

### Create 2010 Data

```{r create-2010-data, eval=F}

# Recreate the Table 1 data as a data frame
raw_2010 <- data.frame(islet = c("Ainsley", "Cooper", "Dudley", "Eastern", "Fern",
                                 "Home", "Leslie", "Lost", "North Fighter", "Portsmouth",
                                 "Sand", "Southern", "Whippoorwill"),
                       area_ha = c(0.1, 110.8, 1.2, 12.7, 0.5, 7.1, 1.7, 1, 0.3,
                                   61.4, 0.8, 8.2, 1.9),
                       percent_cocos = c(1, 0.3, 0, 0.01, 1, 0.43, 0, 0, 0.08,
                                         0.09, 0.13, 0.75, 0.38))

# Save to raw folder
write_csv(raw_2010, file.path(dir_goal, "raw/raw_2010.csv"))
```

### Find Area 

```{r find-2010-area, eval=F}

raw_2010 <- read_csv(file.path(dir_goal, "raw/raw_2010.csv"))

area_2010 <- raw_2010 %>% 
  mutate(area_km = area_ha * 0.01,
         area_cocos = area_km * percent_cocos, 
         region_id = 1) %>% 
  group_by(region_id) %>% 
  summarize(tot_area_cocos = sum(area_cocos),
            total_area = sum(area_km)) %>% 
  mutate(area_non_veg = total_area * 0.0572,
         area_grass = total_area * 0.0759,
         area_forest = total_area - (area_non_veg + area_grass + tot_area_cocos),
         year = 2010) %>% 
  dplyr::select(region_id, year, area_forest)

# Save csv file 
write_csv(area_2010, file.path(dir_goal, "int/trend_area_2010.csv"))
```


## 2016 Data 

### Get Vegetation Community Maps

For habitats we are interested in those dominated by coconut palms and all other forest areas. The following vegetation communities are grouped together and considered "cocos" because they have the same National Vegetation Classification id:  

 - Cocos: Monocultural coconut (Cocos nucifera; greater than 90 percent) dominated forest/woodland    
 - Cocos/Heliotropium: Coconut forest/woodland with abundant *Heliotropium foertherianum* in the canopy and subcanopy         
 
Based on conversations with Alex Wegmann of The Nature Conservancy, the following communities are grouped together to form the other "forest" area:  

 - Heliotropium-Scaevola: Woodland/scrubland dominated by Heliotropium with lesser amounts of Scaevola   
 - Hibiscus: Monoculture stands of Hibiscus shrubs (greater than 75 percent)       
 - Pandanus: Closed or open forests/woodlands with abundant Pandanus 
 - Pisonia: Forest with abundant *Pisonia grandis*          
 - Pisonia/Heliotropium: Forest/woodlands with abundant Pisonia, and an understory of Heliotropium     
 - Scaevola-Heliotropium: Scrublands dominated by Scaevola with lesser amounts of Heloitropium     
 - Terminalia: Woodlands dominated by Terminalia, generally with an open understory   

These community descriptions were taken from the associated [USGS metadata](https://www.sciencebase.gov/catalog/file/get/5c66fc5be4b0fe48cb3aca49?f=__disk__9f%2F2a%2F87%2F9f2a870f06866871d04667b3f6c906896e7ccead&transform=1&allowOpen=true#Entity%20and%20Attribute%20Information).

The following communities are excluded from the analysis:    
 - Lepturus: grassland community not a forest community     
 - Lepturus-Fimbristylis: grassland community not a forest community 
 - Bare ground: Not associated with a vegetation community    
 - Runway (airplane runway): Not associated with a vegetation community     
 - Water (standing water on the islands): Not associated with a vegetation community      

```{r get-2016-vegetation-maps, eval=F}

veg_map_2016 <- st_read(file.path(dir_anx, '_raw_data/USGS_data/Vegetation_Commmunities_Palmyra_Atoll_2016'),
                   layer = "Vegetation_Commmunities_Palmyra_Atoll_2016")

plot(veg_map_2016)

# Shows vegetation communities across all the islands in the atoll along with area in hectares
# group by community and calculate the total area of each type of community

community_areas_2016 <- veg_map_2016 %>% 
  group_by(Community) %>% 
  summarize(comm_area_ha = sum(Area_ha)) 

plot(community_areas)

# Find the percent non-vegetated land (bare ground, water, and runway)
area_non_veg <- community_areas_2016$comm_area_ha[community_areas$Community == "Bare ground"] +
                community_areas_2016$comm_area_ha[community_areas$Community == "Water"] +
                community_areas_2016$comm_area_ha[community_areas$Community == "Runway"]

percent_non_veg <- area_non_veg / sum(community_areas$comm_area_ha) # 4.69%

# Interested in the amount dominated by palm vs the amount dominated by other tree species 
# Find area in square kilometers
area_2016 <- community_areas_2016 %>% 
  dplyr::filter(Community != "Bare ground" & Community != "Water" & Community != "Runway" &
                Community != "Lepturus" & Community != "Lepturus-Fimbristylis") %>% 
  mutate(veg_comm = ifelse(Community == "Cocos" | Community == "Cocos/Heliotropium", "cocos", "forest")) %>% 
  group_by(veg_comm) %>% 
  summarize(area_ha = sum(comm_area_ha)) %>% 
  mutate(area_km = area_ha * 0.01) %>% 
  filter(veg_comm == "forest") %>% 
  rename(area_forest = area_km) %>% 
  mutate(region_id = 1,
         year = 2016) %>% 
  dplyr::select(region_id, year, area_forest)

# Convert to data frame and remove geometry
area_2016 <- as.data.frame(area_2016) %>% 
  dplyr::select(-geometry)

# Save to int folder
write_csv(area_2016, file.path(dir_goal, "int/trend_area_2016.csv"))
```

### Grassland Area 

Find the estimated percent of area occupied by grasslands (Lepturus and Lepturus-Fimbristylis communities). This area will be averaged with the grassland area in 2005 to provide an estimated percent area of grasslands for 2010.  

```{r grassland-area-2016, eval=F}
veg_map_2016 <- st_read(file.path(dir_anx, '_raw_data/USGS_data/Vegetation_Commmunities_Palmyra_Atoll_2016'),
                   layer = "Vegetation_Commmunities_Palmyra_Atoll_2016")

# Find grassland area
grass_area_2016 <- veg_map_2016 %>% 
  filter(Community == "Lepturus" | Community == "Lepturus-Fimbristylis") %>% 
  mutate(comm = "grassland") %>% 
  group_by(comm) %>% 
  summarize(area = sum(Area_ha))

# Find total area
total_area_2016 <- sum(veg_map_2016$Area_ha)

# Percent grassland area
pct_grass_2016 <- grass_area_2016$area / total_area_2016 # 6.40%
```


## 2019 Data

The 2019 area was calculated as part of the `rainforest_data_prep.Rmd`  

```{r get-2019-data, eval=F}

veg_2019 <- read_csv(file.path(dir_goal, "int/veg_area.csv"))

area_2019 <- veg_2019 %>% 
  filter(category == "forest") %>% 
  rename(area_forest = cover_area) %>% 
  mutate(region_id = 1,
         year = 2019) %>% 
  dplyr::select(region_id, year, area_forest)

# Save to int folder
write_csv(area_2019, file.path(dir_goal, "int/trend_area_2019.csv"))
```


## Combine Trend Data

```{r combine-data, eval=F}

# Read in yearly files
area_2005 <- read_csv(file.path(dir_goal, "int/trend_area_2005.csv"))
area_2010 <- read_csv(file.path(dir_goal, "int/trend_area_2010.csv"))
area_2016 <- read_csv(file.path(dir_goal, "int/trend_area_2016.csv"))
area_2019 <- read_csv(file.path(dir_goal, "int/trend_area_2019.csv")) 

# Combine data
trend_area <- area_2005 %>% 
  rbind(area_2010) %>% 
  rbind(area_2016) %>% 
  rbind(area_2019)

# Save to int folder 
write_csv(trend_area, file.path(dir_goal, "int/combined_trend_area.csv"))
```


## Calculate Trend 

Trend is based on a linear model to calculate the rate of change in forest area per year using data from 2010-2019. The trend is the five year average change in rainforest area.      

```{r linear-model, eval=F}

trend_area <- read_csv(file.path(dir_goal, "int/combined_trend_area.csv"))

# Linear model to find change in forest area
trend_regr   <- lm(area_forest ~ year, data = trend_area)
trend_coeffs <- summary(trend_regr)

# Find the rate of change in forest area 
slope <- trend_coeffs$coefficients[2] %>% round(5)

# Find the 5 year trend by multiplying the change per year by five
trend <- slope * 5

# Create the final trend data frame 
rainforest_trend <- data.frame(
  region_id = 1, 
  habitat = "rainforest",
  year = 2019,
  trend = trend
)

# Save trend to output folder
write_csv(rainforest_trend, file.path(dir_goal, "output/rainforest_trend.csv"))
```


## Save to Toolbox

```{r save-to-toolbox, eval=F}
write_csv(rainforest_trend, file.path(dir_scores, 'layers/hab_rainforest_trend.csv'))
```
