---
title: 'OHI+ Palmyra 2020: Habitats Subgoal Rainforest Trend'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
     #in_header: '../../../workflow/templates/ohi_hdr.html' - need to update this
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

# Summary 

This script calculates the trend in rainforest area in 2001, 2010, and 2016 relative to the 2001 potential rainforest area.   

# Data 

Trend comes from three sources of satellite images and published papers:  

2001 Reference Point: The 2001 reference point comes from a satellite image of the island that is discussed in the 2011 Terrestrial Forest Management Plan.     
 - Hathaway, S.A., McEachern K., & Fisher, R.N. (2011). "Terrestrial Forest Management Plan for Palmyra Atoll". U.S. Geological Survey Open-File Report 2011-1007.  
  [Available here](https://permanent.fdlp.gov/gpo119875/ofr20111007.pdf)  
 
2010: The 2010 data is estimated from a study on invasive rats that quantified the percent of land area on each islet occupied by *Cocos nucifera* (Table 1).    
 - Lafferty, K.D. et al. (2010). "Stomach nematodes (*Mastophorus muris*) in rats (*Rattus rattus*) are associated with coconut (*Cocos nucifera*) habitat at Palmyra Atoll. *Journal of Parasitology* 96(1): 16-20.   
  [Available here](https://bioone.org/journals/Journal-of-Parasitology/volume-96/issue-1/GE-2180.1/Stomach-Nematodes-Mastophorus-muris-in-Rats-Rattus-rattus-Are-Associated/10.1645/GE-2180.1.full)   
  
2016: The 2016 data is estimated using a U.S. Geological Survey vegetation community mapping project for Palmyra Atoll. Vegetation maps were completed in October 2016 and published in June 2019. Data was downloaded on July 2, 2020 from the USGS data portal, [here](https://www.sciencebase.gov/catalog/item/5c66fc5be4b0fe48cb3aca49).   
 - Struckhoff, M.A., 2019, Woody species crown and vegetation community data, Palmyra Atoll, 2016: U.S. Geological Survey data release, https://doi.org/10.5066/P9GMPVIO.  
 
 2019: The 2019 data set comes from Table A1 in the Burnett et al. paper   
  - Burnett, M.W. et al. (2019). "Quantifying coconut palm extent on Pacific islands using spectral textural analysis of very high resolution imagery". *International Journal of Remote Sensing* 40(19): 7329-7355   
  [Available here](https://www.tandfonline.com/doi/epub/10.1080/01431161.2019.1594440?needAccess=true). 

# Setup 

```{r setup, message = F, warning = F}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(sf)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_goal   <- '~/github/pal-prep/prep/bd/hab/v2020'
dir_anx    <- file.path(dir_M, 'git-annex/foursite/palmyra')
```


# Methods 

## 2001 Data

The historic percent coverage of palm and rainforest will be used as the reference point for the status score. Historic values are based on a 2001 satellite image evaluated in the U.S. Geological Survey 2011 Terrestrial Forest Management Plan for Palmyra Atoll. There is no downloadable data, these numbers are pulled from the report to create the data.    

This report states that references to the size of the island vary, so a total land area value was used from 2000, which is the closest estimate of the area of the island when the 2001 satellite image was taken. The 2000 estimated land area was manually derived and cited in the 2019 study where the current status data was downloaded.  

The report gives the following vegetation categories and percent coverage:  

 - Cocos: 43%    
 - Scaevola: 29.5%   
 - Pisonia or native forest: 12%   
 - Broadleaf forest (combined with native forest): 6.2%   
 - Hau (combined with native forest): 0.5% 
 
The remaining 8.8% of the land area will be assumed to be non-vegetated.  

Historic data was created into a data frame in the `rainforest_data_prep.Rmd` and saved as `historic_area.csv`.        

```{r 2001-area, eval=F}

raw_2001 <- read_csv(file.path(dir_goal, "int/historic_area.csv"))

# All data should be region_id, year, and area_forest
area_2001 <- raw_2001 %>% 
  filter(category == "forest") %>% 
  rename("area_forest" = "cover_area") %>% 
  mutate(region_id = 1, 
         year = 2001) %>% 
  dplyr::select(region_id, year, area_forest)

# Save to int folder
write_csv(area_2001, file.path(dir_goal, "int/trend_area_2001.csv"))
```


## 2010 Data 

Data for 2010 is recreated from Table 1 in the Lafferty et al paper. The table gives each islet, the islet area (ha), and the percent cocos. Since no specific percent area of non-vegetated land is included in this study, we will use the average percent between the 2001 and 2016 data. The 2001 estimate for non-vegetated land is 8.8% and the 2016 estimate of non-vegetated land is 4.69%. For the 2010 data, we'll assume 6.75% of the land area is non-vegetated and the remainder is considered "forest".    

### Create 2010 Data

```{r create-2010-data, eval=F}

# Recreate the Table 1 data as a data frame
raw_2010 <- data.frame(islet = c("Ainsley", "Cooper", "Dudley", "Eastern", "Fern",
                                 "Home", "Leslie", "Lost", "North Fighter", "Portsmouth",
                                 "Sand", "Southern", "Whippoorwill"),
                       area_ha = c(0.1, 110.8, 1.2, 12.7, 0.5, 7.1, 1.7, 1, 0.3,
                                   61.4, 0.8, 8.2, 1.9),
                       percent_cocos = c(1, 0.3, 0, 0.01, 1, 0.43, 0, 0, 0.08,
                                         0.09, 0.13, 0.75, 0.38))

# Save to raw folder
write_csv(raw_2010, file.path(dir_goal, "raw/raw_2010.csv"))
```

### Find Area 

```{r find-2010-area, eval=F}

raw_2010 <- read_csv(file.path(dir_goal, "raw/raw_2010.csv"))

area_2010 <- raw_2010 %>% 
  mutate(area_km = area_ha * 0.01,
         area_cocos = area_km * percent_cocos, 
         region_id = 1) %>% 
  group_by(region_id) %>% 
  summarize(tot_area_cocos = sum(area_cocos),
            total_area = sum(area_km)) %>% 
  mutate(area_non_veg = total_area * 0.0675,
         area_forest = total_area - (area_non_veg + tot_area_cocos),
         year = 2010) %>% 
  dplyr::select(region_id, year, area_forest)

# Save csv file 
write_csv(area_2010, file.path(dir_goal, "int/trend_area_2010.csv"))
```


## 2016 Data 

### Get Vegetation Community Maps

For habitats we are interested in those dominated by coconut palms and all other forest areas. The following vegetation communities are grouped together and considered "cocos" because they have the same National Vegetation Classification id:  

 - Cocos: Monocultural coconut (Cocos nucifera; greater than 90 percent) dominated forest/woodland    
 - Cocos/Heliotropium: Coconut forest/woodland with abundant *Heliotropium foertherianum* in the canopy and subcanopy         
 
The following communities are grouped together to form the other "forest" area:  

 - Heliotropium-Scaevola: Woodland/scrubland dominated by Heliotropium with lesser amounts of Scaevola; both species are native to Palmyra   
 - Hibiscus: Monoculture stands of Hibiscus shrubs (greater than 75 percent), *Hibiscus tiliaceus* is nonnative      
 - Lepturus: Herbaceous community dominated by Pacific Island thintail (*Lepturus repens*); native to Palmyra     
 - Lepturus-Fimbristylis: Mowed areas dominated by Pacific Island thintail (Lepturus repens) and/or Fimbristylis; both species are native to Palmyra   
 - Pandanus: Closed or open forests/woodlands with abundant Pandanus (native), often with Cocos codominant 
 - Pisonia: Forest with abundant *Pisonia grandis* (native), sometimes co-dominant with or subordinate to coconut       
 - Pisonia/Heliotropium: Forest/woodlands with abundant Pisonia, and an understory of Heliotropium; both species are native to Palmyra      
 - Scaevola-Heliotropium: Scrublands dominated by Scaevola with lesser amounts of Heloitropium; both species are native to Palmyra   
 - Terminalia: Woodlands dominated by Terminalia, generally with an open understory; *Terminalia catappa* is generally regarded as nonnative invasive  

These community descriptions were taken from the associated [USGS metadata](https://www.sciencebase.gov/catalog/file/get/5c66fc5be4b0fe48cb3aca49?f=__disk__9f%2F2a%2F87%2F9f2a870f06866871d04667b3f6c906896e7ccead&transform=1&allowOpen=true#Entity%20and%20Attribute%20Information) and species were compared against a list of native vascular plant species for Palmyra, found [here](http://www.hawaii.edu/gk-12/palmyra/specieslist/nativeplants.htm). *Heliotropium foertherianum* is also known as *Tournefortia argentea*, which is native to Palmyra.      

Three communities are filtered out because they are not dominated by trees or vegetation of any kind:  

 - Bare ground   
 - Runway (airplane runway)    
 - Water (standing water on the islands)   

```{r get-2016-vegetation-maps, eval=F}

veg_map_2016 <- st_read(file.path(dir_anx, '_raw_data/USGS_data/Vegetation_Commmunities_Palmyra_Atoll_2016'),
                   layer = "Vegetation_Commmunities_Palmyra_Atoll_2016")

plot(veg_map_2016)

# Shows vegetation communities across all the islands in the atoll along with area in hectares
# group by community and calculate the total area of each type of community

community_areas_2016 <- veg_map_2016 %>% 
  group_by(Community) %>% 
  summarize(comm_area_ha = sum(Area_ha)) 

plot(community_areas)

# Find the percent non-vegetated land (bare ground, water, and runway)
area_non_veg <- community_areas_2016$comm_area_ha[community_areas$Community == "Bare ground"] +
                community_areas_2016$comm_area_ha[community_areas$Community == "Water"] +
                community_areas_2016$comm_area_ha[community_areas$Community == "Runway"]

percent_non_veg <- area_non_veg / sum(community_areas$comm_area_ha) # 4.69%

# Interested in the amount dominated by palm vs the amount dominated by other tree species 
# Find area in square kilometers
area_2016 <- community_areas_2016 %>% 
  dplyr::filter(Community != "Bare ground" & Community != "Water" & Community != "Runway") %>% 
  mutate(veg_comm = ifelse(Community == "Cocos" | Community == "Cocos/Heliotropium", "cocos", "forest")) %>% 
  group_by(veg_comm) %>% 
  summarize(area_ha = sum(comm_area_ha)) %>% 
  mutate(area_km = area_ha * 0.01) %>% 
  filter(veg_comm == "forest") %>% 
  rename(area_forest = area_km) %>% 
  mutate(region_id = 1,
         year = 2016) %>% 
  dplyr::select(region_id, year, area_forest)

# Convert to data frame and remove geometry
area_2016 <- as.data.frame(area_2016) %>% 
  dplyr::select(-geometry)

# Save to int folder
write_csv(area_2016, file.path(dir_goal, "int/trend_area_2016.csv"))
```

## 2019 Data

The 2019 area was calculated as part of the `rainforest_data_prep.Rmd`  

```{r get-2019-data, eval=F}

veg_2019 <- read_csv(file.path(dir_goal, "int/veg_area.csv"))

area_2019 <- veg_2019 %>% 
  filter(category == "forest") %>% 
  rename(area_forest = cover_area) %>% 
  mutate(region_id = 1,
         year = 2019) %>% 
  dplyr::select(region_id, year, area_forest)

# Save to int folder
write_csv(area_2019, file.path(dir_goal, "int/trend_area_2019.csv"))
```


## Combine Trend Data

```{r combine-data, eval=F}

# Read in yearly files
area_2010 <- read_csv(file.path(dir_goal, "int/trend_area_2010.csv"))
area_2016 <- read_csv(file.path(dir_goal, "int/trend_area_2016.csv"))
area_2019 <- read_csv(file.path(dir_goal, "int/trend_area_2019.csv")) 

# Combine data
trend_area <- area_2010 %>% 
  rbind(area_2016) %>% 
  rbind(area_2019)

# Save to int folder 
write_csv(trend_area, file.path(dir_goal, "int/combined_trend_area.csv"))
```


## Calculate Trend 

Trend is based on a linear model to calculate the rate of change in forest area per year using data from 2010-2019. The trend is the five year average change in rainforest area.      

```{r linear-model, eval=F}

trend_area <- read_csv(file.path(dir_goal, "int/combined_trend_area.csv"))

# Linear model to find change in forest area
trend_regr   <- lm(area_forest ~ year, data = trend_area)
trend_coeffs <- summary(trend_regr)

# Find the rate of change in forest area 
slope <- trend_coeffs$coefficients[2] %>% round(5)

# Find the 5 year trend by multiplying the change per year by five
trend <- slope * 5

# Create the final trend data frame 
rainforest_trend <- data.frame(
  region_id = 1, 
  habitat = "rainforest",
  year = 2019,
  trend = trend
)
```


## Save to Toolbox

```{r save-to-toolbox, eval=F}
# save to toolbox
dir_scores <- '~/github/pal-scores/region'

write_csv(rainforest_trend, file.path(dir_scores, 'layers/hab_rainforest_trend.csv'))
```
