---
title: 'Habitat Destruction Pressure'
subtitle: 'OHI+ Palmyra 2020 Assessment'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3 
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     in_header: '~/github/pal-prep/src/templates/pal_hdr.html' 
  pdf_document:
    toc: true
---

# Summary 

This script calculates the pressures from habitat destruction on Palmyra Atoll. Habitat destruction typically includes activities like deforestation, dredging, bottom trawling etc. None of these activities take place on Palmyra as the island and offshore waters are fully protected. Therefore, the pressure from habitat destruction is low.     

# Data 

Data to inform this layer was obtained from verbal communications with Alex Wegmann at The Nature Conservancy on July 31, 2020.  

# Setup 

```{r setup, message=F, warning=F, results='hide'}

library(tidyverse)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_prs   <- '~/github/pal-prep/prep/pressures/hd/v2020'
```

# Methods 

## Pressure Score 

Given that there are no activities contributing to habitat destruction permitted on the island or in the offshore area, this pressure score is zero. However, we still want to include this layer as part of the overall calculation. We will create a layer with region_id, year, pressure_score (0) to save in the toolbox.   

```{r pressure-score, eval=F}
# Years
years <- 2014:2020

# Pressure score 
hd_pressure <- data.frame(
  region_id = rep(1, length(years)),
  year = years,
  pressure_score = rep(0, length(years))
)

# Save to output file 
write_csv(hd_pressure, file.path(dir_prs, "output/hd_pressure.csv"))
```

## Save to Toolbox 

```{r save-to-toolbox, eval=F}
hd_pressure <- read_csv(file.path(dir_prs, "output/hd_pressure.csv"))
write_csv(hd_pressure, file.path(dir_scores, "layers/prs_hd.csv"))
```

