---
title: "sst"
output: html_document
---

# Summary
For the sea surface temperature pressure we calculated the number of times Palmyra's weekly SST was greater than the climatological mean for that week (an anomaly: greater than mean + one standard deviation) and summed the number of weekly anomalies in a single year. The maximum value a cell could have is 52 which would mean that cell had anomalous SST temperatures for each week of the year.

To account for annual variation, we look at Sea Surface Temperature anomalies in 5 year periods, so the maximum value possible per cell is 260 anomalous weeks. To rescale the values from 0 to 1 we set a reference point. Previously, the reference point for SST has just been the maximum difference in anomalous weeks between the most recent time period and a historical reference period (1985-1989).

This time we have decided to use a reference point that represents a regime shift. Once a given cell is anomalous for more than 50% of a five-year period, it has shifted into a new regime. All cells that have a value greater than 130 weeks (51% of a 5 year time period) are assigned a 1. The rest of the cells are scaled to this reference point by dividing by 130.

# Data 

The sea surface temperature layer uses data from [NOAA's Coral Health and Monitoring Program](https://www.coral.noaa.gov/champportal/).   

**Version**: v1.12, released September 12, 2018
**Values**: Sea surface temperature (Celsius)      
**Time Range**: Daily data from January 1, 1998 - December 31, 2019     
**Format**: csv  

Data were downloaded on July 22, 2020.  

Citation: National Oceanic and Atmospheric Administration (2018). Coral Health and Monitoring Program v1.12. Accessed [June 22, 2020].   

# Setup 

```{r setup, message = F, warning = F}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(janitor)
library(lubridate)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_prs   <- '~/github/pal-prep/prep/pressures/sst/v2020'
```

# Methods

## Raw Data 

There are two sensors collecting temperature data on Palmyra Atoll. We will average the readings from both sensors to get a single temperature value for each day from 1998 - 2019.  

```{r get-raw-data, eval=F}
temp_raw <- read_csv(file.path(dir_anx, "_raw_data/NOAA_data/champPortal-2020-07-22-1832-r089.csv")) %>% 
  clean_names()

# Wrangle dates and find the average of temperature readings
temp_data <- temp_raw %>% 
  mutate(date = as.Date(temp_raw$datetime),
         year = year(date),
         avg_temp = (pacp1_oisst_mwsst_degc + ppst1_oisst_mwsst_degc) / 2,
         obs_id = seq(1:nrow(temp_raw))) %>% 
  dplyr::select(obs_id, date, year, avg_temp)

## Assign first 7 days to Week 1, next 7 days to Week 2 etc. 
# Group by year and each year should have 52 weeks, plus one extra day, leap years will have 2 extra days
wk <- c(rep(1:52, each = 7), 53)
wk_leap <- c(rep(1:52, each = 7), 53, 53)

# Define which years were leap years from 1998-2019
leap_yrs <- c(2000,2004,2008,2012,2016)

# Add days of the week and the week number
temp_dates <- temp_data %>% 
  group_by(year) %>% 
  mutate(wk = ifelse(year %in% leap_yrs, wk_leap, wk))


# Save to int folder
write_csv(temp_dates, file.path(dir_prs, "int/sst_daily.csv"))
```

## Weekly Average Temperatures

Using the daily temperature data, find the mean weekly temperature and weekly temperature standard deviation.

```{r weekly-anomalies, eval=F}
sst_daily <- read_csv(file.path(dir_prs, "int/sst_daily.csv"))

# Find weekly mean and standard deviation 
sst_wk <- sst_daily %>% 
  group_by(year, wk) %>% 
  mutate(wk_avg_temp = mean(avg_temp),
         wk_sd_temp = sd(avg_temp)) %>% 
  ungroup() %>% 
  dplyr::select(obs_id, year, wk, wk_avg_temp, wk_sd_temp)

# Save to int folder
write_csv(sst_wk, file.path(dir_prs, "int/sst_wkly_avg.csv"))
```

