---
title: 'OHI+ Palmyra 2020 - Iconic species subgoal'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
     #in_header: '../../../workflow/templates/ohi_hdr.html' - need to update this
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

# Summary:  Iconic Species Subgoal (Sense of Place)

This script prepares scores (status and trend) for Iconic Species in Palmyra.  For each iconic marine species, extinction risk category is pulled based on current and past assessments; by tracking the assessed extinction risk over time, we can understand the trends of extinction risk for iconic species directly rather than using the "population trend" method from prior OHI assessments.

The Iconic Species sub-goal model calculates status based upon an unweighted average of species health for all 'iconic' species found within the reporting region.

From Halpern et al (2012):

> Iconic species are those that are relevant to local cultural identity through a species’ relationship to one or more of the following: 1) traditional activities such as fishing, hunting or commerce; 2) local ethnic or religious practices; 3) existence value; and 4) locally-recognized aesthetic value (e.g., touristic attractions/common subjects for art such as whales). Habitat-forming species are not included in this definition of iconic species, nor are species that are harvested solely for economic or utilitarian purposes (even though they may be iconic to a sector or individual). ...
> Ultimately, almost any species can be iconic to someone, and so the intent with this goal was to focus on those species widely seen as iconic within a country, and iconic from a cultural or existence value (rather than for a livelihoods or extractive reason). ...
> The reference point is to have the risk status of all assessed species as Least Concern (i.e., a goal score = 1.0)
The Status of this sub-goal (X~ICO~) is then the % of iconic species in each threat category (as defined by the IUCN Red List), such that:

$$X_{ICO} = \frac{\displaystyle\sum_{category}S_{cat}*w_{cat}}{\displaystyle\sum_{category}S_{cat}}$$
where for each IUCN threat category:
* *S~cat~* is the number of assessed species in the category
* *w~cat~* is the status weight assigned for that category (note, these are the inverse of the risk value used in the SPP calculations):
    * 'LC' = 1.0, 'NT' = 0.8, 'VU' = 0.6, 'EN' = 0.4, 'CR' = 0.2, 'EX' = 0.0

ICO trend is calculated in a similar manner, but weightings are assigned according to IUCN population trend: 'Decreasing' = -0.5, 'Stable' = 0.0, 'Increasing' = +0.5.  


# Data Sources

**List of iconic species:**

This initial list of iconic species is compiled based on USFWS birds, mammals, reptiles, and invertebrates list.

* __Reference__: 
    * U.S. Fish and Wildlife Service. (2016, January 14). Wildlife and Habitat. Retrieved from <https://www.fws.gov/refuge/Palmyra_Atoll/wildlife_and_habitat/index.html> 
        * __Accessed__: 04 June 2020   
        
Birds: brown booby (Sula leucogaster), masked booby (Sula dactylatra), red-footed booby (Sula sula), black noddy (Anous minutus),
brown noddy (Anous stolidus), great frigatebird (Fregata minor),sooty tern (Onychoprion fuscata),
bristle-thighed curlew (Numenius tahitiensis), pacific golden plover (Pluvialis fulva), wandering tattler (Heteroscelus incanus),
ruddy turnstone (Arenaria interpres)   

Mammals: pacific bottlenose dolphins (Tursiops aduncus), spinner dolphins (Stenella longirostris),
melon-headed whales (Peponocephala electra)   

Reptiles: green turtles (Chelonia mydas), hawksbill turtles (Eretmochelys imbricata)    

Invertebrates: stony corals (many species), giant clams (Tridacna gigas), coconut crabs (Birgus latro)   

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rredlist)
library(tidyverse)
library(dplyr)
library(raster)
library(here)

source(here('src/R/common.R'))

api_file <- file.path(dir_M, 'git-annex/globalprep/spp', 'api_key_gc.csv')
api_key <- scan(api_file, what = 'character')
```

Get US Minor Outlying Islands species 
```{r}
spp_country_url <- 'http://apiv3.iucnredlist.org/api/v3/country/getspecies/id/%s?token=%s'

countries <- rl_countries(key = api_key)$results

#UM is United States Minor Outlying Islands country code
# UM species risks (1532 species)
spp_um <- rl_sp_country('UM', key = api_key)$result %>%
  mutate(country_code = "UM") %>%
  rename("iucn_sid" = "taxonid")

# Create vector of iconic species from the USFWS webpage to cross reference (17 spp total)
# Note this list excludes specific coral and fish species that weren't listed on the USFWS webpage
ico <- c("Sula leucogaster", "Sula dactylatra", "Sula sula", "Anous minutus", "Anous stolidus",
          "Fregata minor", "Onychoprion fuscata", "Numenius tahitiensis", "Pluvialis fulva",
          "Heteroscelus incanus", "Arenaria interpres", "Tursiops aduncus", "Stenella longirostris",
          "Chelonia mydas", "Eretmochelys imbricata", "Tridacna gigas", "Birgus latro")

# Filter the UM spp data for Palmyra iconic spp 
pal_ico <- spp_um %>% 
  filter(scientific_name %in% ico) # 11 matching the ico spp list
# list all birds (except tern and tattler) plus spinner dolphins and green turtles
```


