---
title: 'OHI+ Palmyra 2020: Invasive Species Management Resilience Layer'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    toc_depth: 3 
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     in_header: '../../../src/templates/ohi_hdr.html' 
  pdf_document:
    toc: true
---

# Summary 

Regulatory resilience is meant to describes the institutional measures (e.g., rules, regulations, and laws) designed to address ecological pressures. This script calculates resilience from the management of invasive species on Palmyra Atoll. 

# Data 

Data to inform the preparation of this layer comes from the "Existing Non-Native Species Laws, Policies, Protocols" section of the 2010 Palmyra Biosecurity Plan.  

Citation: Hathaway, S.A. and Fisher, R.N. (2010). Biosecurity Plan for Palmyra Atoll: U.S. Geological Survey Open-File Report 2010-1097, 80p.  
   Available [here](https://pubs.er.usgs.gov/publication/ofr20101097)   


# Setup 

```{r setup, message = F, warning = F}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_res   <- '~/github/pal-prep/prep/resilience/ecological/invasive_species/v2020'
```


# Methods 

## Resilience Score 

A resilience layer is created for management of invasive species based on three criteria:  

 - Is there a management plan in place?  
 - Is there monitoring and assessment?   
 - Is the management plan effective?  

A yes will be assigned a value of 1, in-development will be assigned a value of 0.5, a no will be assigned a value of 0. The Palmyra Atoll Biosecurity Plan established in 2010 details a series of adaptive management plans designed to achieve TNC’s mission toward the protection and enhancement of native wildlife and habitat. Since the established of the plan, major actions to remove priority species, such as invasive rats (*Rattus rattus*) have been undertaken. Monitoring for invasive species continues to take place and there are additional plans to continue removal of other high priority species like coconut palm (*Cocos nucifera*).   
 
| Category         | Response         | Score | 
|------------------|------------------|-------|
| Management Plan? | Yes              | 1     |
| Monitoring?      | Yes              | 1     |
| Effective?       | Yes              | 1     | 

```{r mgmt-resilience-score, eval=F}

years <- 2014:2020

sp_mgmt <- data.frame(
  region_id = rep(1, length(years)),
  year = years,
  resilience_score = rep(1, length(years))
)

# Save to output folder
write_csv(sp_mgmt, file.path(dir_res, "output/sp_mgmt.csv"))
```

## Save to Toolbox

```{r save-to-toolbox, eval=F}
sp_mgmt <- read_csv(file.path(dir_res, "output/sp_mgmt.csv"))
write_csv(sp_mgmt, file.path(dir_scores, "layers/res_sp_mgmt.csv"))
```

