---
title: 'Resource Access Opportunities'
subtitle: 'OHI+ Palmyra 2020 Assessment'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3 
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     in_header: '~/github/pal-prep/src/templates/pal_hdr.html' 
  pdf_document:
    toc: true
---

# Summary 

This script generates a temporary layer to use in place of a status score for the Resource Access Opportunities goal. 

# Data

Data for this layer is supposed to come from the fishing log on Palmyra Atoll. The log was not digitized at the time of the assessment and therefore couldn't be used. 

# Setup

```{r setup, message=F, warning=F, results='hide'}

library(tidyverse)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_goal   <- '~/github/pal-prep/prep/rao/v2020'
```


# Methods

The toolbox requires the layer to have a column for region_id, year, and status (a score 0-1). For trend we like to have five years of data, so the years will span 2014-2020, but since the data is not available to complete this layer in time for the assessment, a placeholder with a value of NA will be used.   

## Create Temporary Layer 

```{r create-temp-layer, eval=F}
# Set years 
years <- 2014:2020

rao_status <- data.frame(
  region_id = rep(1, length(years)), 
  year = years, 
  status = rep(NA, length(years))
)

# Save placeholder to output
write_csv(rao_status, file.path(dir_goal, "output/rao_status_placeholder.csv"))
```

## Save to Toolbox

```{r save-to-toolbox, eval=F, message=F, warning=F}
rao_status <- read_csv(file.path(dir_goal, "output/rao_status_placeholder.csv"))
write_csv(rao_status, file.path(dir_scores, "layers/rao_status_placeholder.csv"))
```

