---
title: 'Resource Access Opportunities'
subtitle: 'OHI+ Palmyra 2020 Assessment'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3 
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     in_header: '~/github/pal-prep/src/templates/pal_hdr.html' 
  pdf_document:
    toc: true
---

# Summary 

This script calculates a status score for access to sustenance fishing opportunities around Palmyra Atoll from 2014 - 2020.   

# Data

Data for this layer is supposed to come from the fishing log on Palmyra Atoll. The log was not digitized at the time of the assessment, but estimates were obtained via email from Stefan Kropidlowski at the U.S. Fish and Wildlife Service on August 18, 2020. 

# Setup

```{r setup, message=F, warning=F, results='hide'}

library(tidyverse)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_goal   <- '~/github/pal-prep/prep/rao/v2020'
```


# Methods

The USFWS estimates about 20 fish are caught each year for sustenance fishing. We will use this estimate for all years from 2014 - 2020. These estimates should be updated once the sustenance fishing log on Palmyra Atoll is digitized.   

## Create Raw Data 

```{r create-raw-data, eval=F}
# Set years 
years <- 2014:2020

catch_raw <- data.frame(
  region_id = rep(1, length(years)), 
  year = years, 
  catch = rep(20, length(years))
)

# Save to raw 
write_csv(catch_raw, file.path(dir_goal, "raw/catch_raw.csv"))
```

## Calculate Status 

**Reference Point:** Up to 20 fish caught for sustenance by residents of Palmyra Atoll per year. 

The current status will be scaled to the reference point using the following equation:      

$status~=~1~-~[(catch_{y}~-catch_{ref,y})~/~catch_{ref,y}]$   

Where $catch_{y}$ is the catch in each year (y) and $catch_{ref,y}$ is the reference point catch in year (y). Any value greater than one will be given a status of one, and any value less than zero will be given a zero. This gives a perfect score to any year where the number of fish caught is at or below the reference point and a zero if the number of fish caught is double or more than double the reference point. This method of scoring is based on The Nature Conservancy's management goals to keep fishing at or below the reference point level and reflective of the view that fishing at more than twice the reference point would cause serious damage to the ecosystem.        

```{r calculate-status, eval=F}

catch <- read_csv(file.path(dir_goal, "raw/catch_raw.csv"))

rao_status <- catch %>% 
  mutate(ref = 20,
         status = 1 - (catch - ref) / ref) %>% 
  mutate(status = ifelse(status > 1, 1, status)) %>% 
  dplyr::select(region_id, year, status)

# Save to output folder
write_csv(rao_status, file.path(dir_goal, "output/rao_status.csv"))
```

## Save to Toolbox

```{r save-to-toolbox, eval=F, message=F, warning=F}
rao_status <- read_csv(file.path(dir_goal, "output/rao_status.csv"))
write_csv(rao_status, file.path(dir_scores, "layers/rao_status.csv"))
```

