---
title: 'OHI+ Palmyra 2020: Habitat Services Subgoal Carbon Storage'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
     #in_header: '../../../workflow/templates/ohi_hdr.html' - need to update this
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

# Summary
This script classifies tree species into native and nonnative groups and finds the percent of total area occupied by native tree species. The carbon storage status is based on the percent of land occupied by native tree species compared to the reference point percentage of land occupied by native tree species.   


# Setup

```{r setup, message = F, warning = F}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(sf)
library(rgdal)
library(lwgeom)

# Source and set file paths
source(here('src/R/common.R'))

dir_goal   <- '~/github/pal-prep/prep/hs/cs/v2020'
dir_anx    <- file.path(dir_M, 'git-annex/foursite/palmyra')
```


# Methods

## Get Tree Cover Maps

Tree cover maps were downloaded from the U.S. Geological Survey vegetation community mapping project for Palmyra Atoll, found [here](https://www.sciencebase.gov/catalog/item/5c66fc5be4b0fe48cb3aca49).    

Data was downloaded on July 7, 2020, and vegetation maps were completed in October 2016 and published in June 2019.  

```{r get-tree-cover-maps, eval=F}

trees <- st_read(file.path(dir_anx, '_raw_data/USGS_data/Woody_Crowns_Palmyra_Atoll_2016'),
                   layer = "Woody_Crowns_Palmyra_Atoll_2016")

# Warning that some polygons don't have four points
# Check geometry validity
validity <- st_is_valid(trees)
trees$valid <- validity

# Find invalid polygons
invalid <- trees %>% 
  dplyr::filter(valid != TRUE) # 16 invalid polygons

# Check reason for invalid geometries
st_is_valid(invalid, reason = TRUE) # All have self-intersection

# Fix self intersections using st_make_valid
make_valid <- sf::st_make_valid(invalid) %>% 
  st_cast("MULTIPOLYGON") # Retain geometry type multipolygon

# Check if it worked
st_is_valid(make_valid) # Get all true now 

# Bind back together
trees_edit <- trees %>% 
  dplyr::filter(trees$valid == TRUE) %>% 
  rbind(make_valid) %>% # Still missing three observations
  dplyr::select(-valid)

# Save to the raw folder 
st_write(trees_edit, file.path(dir_goal, 'raw/tree_map.shp'), driver = "ESRI Shapefile")

# Check for NA valid geometries instead of False
# Not sure what to do about these 
geom_na <- trees %>% 
  dplyr::filter(is.na(valid)) # Found missing ones - these geometries are corrupt not invalid

```

## Categorize Native and Nonnative Species

The carbon storage goal is based on percent coverage of native tree species, we need to identify which of the mapped species are native vs. nonnative.   

The following tree species are included in the mapping project and we have designated if they are native or nonnative. Category designations were assigned based species lists found in Tables 1 and 2 of the U.S. Geological Survey 2011 Terrestrial Forest Management Plan for Palmyra Atoll, found [here](https://permanent.fdlp.gov/gpo119875/ofr20111007.pdf).         

 - *Araucaria*, species unknown: Nonnative      
 - *Artocarpus altilis*: Nonnative   
 - *Barringtonia asiatica*: Native   
 - *Calophyllum inophyllum*: Nonnative   
 - *Casuarina equisetifolia*: Nonnative     
 - *Coccoloba uvifera*: Nonnative   
 - *Cocos nucifera*: Nonnative   
 - *Cordia subcordata*: Native   
 - *Guettarda speciosa*: Native   
 - *Heliotropium foertherianum* also known as *Tournefortia argentea*: Native   
 - *Hernandia nymphaeifolia*: Native (not on the lists but *Hernandia sonora* is native)    
 - *Hibiscus tiliaceus*: Nonnative   
 - *Ipomoea*: Native   
 - *Nephrolepis*: Nonnative   
 - *Pandanus tectorius* updated from *Pandanus fischerianus*: Native  
 - *Phymatosorus scolopendria*: Native 
 - *Pisonia grandis*: Native   
 - *Premna serratifolia*: Native   
 - *Scaevola taccada*: Native   
 - *Terminalia catappa*: Nonnative   
 - Unknown: excluded from analysis    
 - Water, standing water: excluded from analysis    
 
 
```{r add-native-status, eval=F}
# Read in the shp with fixed geometries 
tree_map <- st_read(file.path(dir_goal, "raw"), layer = "tree_map")

# List of native plants:
natives <- c("Barringtonia asiatica", "Cordia subcordata", "Guettarda speciosa",
             "Heliotropium foertherianum", "Hernandia nymphaeifolia", "Ipomoea",
             "Pandanus tectorius", "Pisonia grandis", "Phymatosorus scolopendria",
             "Premna serratifolia", "Scaevola taccada")

# Add status:
tree_cat <- tree_map %>% 
  mutate(category = ifelse(species %in% natives, "native", "nonnative"))

# Save to int folder
write_sf(tree_cat, file.path(dir_goal, 'int/tree_category.shp'), driver = "ESRI Shapefile")
```
 
## Calculate Area 

Find the area of native vs. nonnative trees on the islands   

```{r calculate-area, eval=F}

tree_cat <- st_read(file.path(dir_goal, "int"), layer = "tree_category") 

# Change area in square meters to area in square km 
# Find area of native and nonnative species
tree_area <- tree_cat %>% 
  mutate(area_km = Area * 0.000001) %>% 
  group_by(category) %>% 
  summarize(geometry = st_union(geometry),
            area_km = sum(area_km))

# Check aggregation 
plot(tree_area)

# Save to int folder
write_sf(tree_area, file.path(dir_goal, "int/tree_area.shp"), driver = "ESRI Shapefile", overwrite = TRUE)

```

## Calculate Status
 
Reference Point: 95% of the forest area dominated by native trees   


The status is calculated as the current percent of forest dominated by native trees over the refernce point percentage   
$status~=~percent~cover~/~reference~point~percent~cover$   
 
```{r calc-status, eval=F}

tree_area <- st_read(file.path(dir_goal, "int"), layer = "tree_area") 

# Define reference point
ref_point <- 0.95 

# Find total area
total_area_km <- sum(tree_area$area_km)

# Find percent area for both native and nonnative species
percent_area <- tree_area %>% 
  mutate(total_area = total_area_km) %>% 
  mutate(percent_area = area_km / total_area)

# Calculate status 
cs_status <- percent_area %>% 
  filter(category == "native") %>% 
  mutate(status = percent_area / ref_point,
         region_id = 1,
         year = 2020) %>% 
  dplyr::select(region_id, year, status)

# Don't need to save the geometries in the output
cs_status <- as.data.frame(cs_status) %>% 
  dplyr::select(-geometry)

# Save as csv to output
write_csv(cs_status, file.path(dir_goal, "output/cs_status.csv"))
```
 
 