---
title: 'Carbon Storage Sub-goal'
subtitle: 'OHI+ Palmyra 2020 Assessment'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3 
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     in_header: '~/github/pal-prep/src/templates/pal_hdr.html' 
  pdf_document:
    toc: true
---

# Summary

This script creates a placeholder layer for the Carbon Storage goal. Little is known about the carbon sequestration rates of atoll rainforests and currently there is not information to calculate a meaningful status score.         

# Data 

Data from this layer is meant to come from tree cover maps from U.S. Geological Survey vegetation community mapping project for Palmyra Atoll, available [here](https://www.sciencebase.gov/catalog/item/5c66fc5be4b0fe48cb3aca49). However, there is not enough accompanying information on carbon sequestration rates to calculate a complete status score.  

# Setup

```{r setup, message=F, warning=F, results='hide'}

library(tidyverse)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_goal   <- '~/github/pal-prep/prep/hs/cs/v2020'
```


# Methods

Since there is not enough information on the carbon sequestration rates of atoll rainforests, we will use a placeholder layer with a status score of 'NA'. There is an ongoing carbon accounting project on Palmyra Atoll which could be used to inform this goal in the future.   

## Create Placeholder Layer

```{r create-placeholder-layer, eval=F}

cs_status <- data.frame(
  region_id = 1, 
  year = 2020, 
  status = NA
)

# Save to output folder
write_csv(cs_status, file.path(dir_goal, "output/cs_status_placeholder.csv"))
```

## Save to Toolbox 

```{r save-to-toolbox, eval=F}
cs_status <- read_csv(file.path(dir_goal, "output/cs_status_placeholder.csv"))
write_csv(cs_status, file.path(dir_scores, 'layers/hs_carbon_storage_status_placeholder.csv'))
```
 
 