---
title: "create_pal_region"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     # in_header: '~/github/ne-prep/src/templates/ohi_hdr.html' # update with own header
  pdf_document:
    toc: true
---

## Summary

This script shows how I created the shapefile for the Palmrya region. I used the following shapefiles:  

 * USFWS Special Area Designations - Marine National Monument Boundary
 * GADM country boundaries

## Data Sources

**Marine National Monument**  

* __Reference__: 
    * USFWS, NWRS, Division of Realty:. 2020. FWS Special Designation Areas - April 2020    <https://ecos.fws.gov/ServCat/Reference/Profile/116244>
        * __Date Created__: 25 February 2016
        * __Date Updated__: 22 April 2020
        * __Accessed__: 5 June 2020   
* __Description__: Lands and waters administered by the USFWS in North America, U.S. Trust Territories and Possessions  
* __Spatial Extent__: West Bounding Longitude: -179.150067, East Bounding Longitude: 179.77173, North Bounding Latitude: 70.329407, South Bounding Latitude: -14.559753
* __Format__: Shapefile  

**Land Boundary**  

* __Reference__: 
    * NOA, Office of Coast Surveys. 2018. Shoreline Palmyra Atoll <https://catalog.data.gov/dataset/shoreline-palmyra-atoll>  
        * __Date Created__: 15 January 2016  
        * __Date Updated__: 31 January 2018  
        * __Metadata Updated__: 27 February 2019  <https://catalog.data.gov/harvest/object/98b61b2b-28e5-4382-bc98-9a655c6c67e1/html>   
        * __Accessed__: 8 June 2020
* __Description__: Shoreline of Palmyra Atoll 
* __Spatial Extent__: West Bounding Longitude: -162.112, East Bounding Longitude: -162.049, North Bounding Latitude: 5.893, South Bounding Latitude: 5.871  
* __Format__: Shapefile  

```{r setup, message = F, warning = F, results = 'hide'}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

source('~/github/pal-prep/src/R/common.R') ### Palmyra common.R

dir_git <- '~/github/pal-prep'
dir_anx <- file.path(dir_M, 'git-annex/foursite/palmyra')

library(tidyverse)
library(sf)
library(mapview)
library(mapedit)
library(leaflet)
```


# Marine National Monument Boundary shapefile

The USFWS shapefile is for all of the U.S. and U.S. territories, we want to clip it to just include the National Marine Monument boundary around Palmyra and Kingman Reef.  

```{r national-monument}
# Read in shapefile for all special areas designations
usfws_sp_area <- st_read(file.path(dir_anx, 'spatial/usfws_special_areas'), layer = "SpecialDesignation")

# Find only the pacific marine national monument palmyra unit and retain the geometry and an 'id' 
marine_monument <- usfws_sp_area %>% 
  filter(CMPXNAME == "MARINE NATIONAL MONUMENTS OF THE PACIFIC") %>% 
  filter(COMMENTS == "KINGMAN REEF & PALMYRA ATOLL UNIT") %>% 
  mutate(id = "marine_monument") %>% 
  dplyr::select(id, geometry) %>% 
  st_transform(mollCRS)

# View to make sure it's correct 
mapview(marine_monument)

# Write as shapefile:
st_write(marine_monument, file.path(dir_git, 'spatial/shapefiles/marine_monument_boundary.shp'), driver = "ESRI Shapefile")
```

# Palmyra shoreline

Check the shapefile for Palmyra's shoreline  

```{r pal-shoreline}
# Read in shapefile for Palmyra shoreline and retain the geometry and an 'id'
shoreline <- st_read(file.path(dir_anx, '_raw_data/PACIOOS_data/pat_ocs_all_shoreline'), layer = "pat_ocs_all_shorelinePolygon") %>% 
  mutate(id = "shoreline") %>% 
  dplyr::select(id, geometry) %>% 
  st_transform(mollCRS) 

# View to make sure it's correct 
mapview(shoreline)

# Write as shapefile:
st_write(shoreline, file.path(dir_git, 'spatial/shapefiles/pal_shoreline.shp'), driver = "ESRI Shapefile")
```


# Create 3 nm buffer

Create a 3 nautical mile buffer around the shoreline. Units for shoreline is in meters 3 nm = 5556 meters  

```{r create-buffer}
# create buffer 
pal_buff <- st_buffer(shoreline, 5556) %>% 
  mutate(id_buff = "buffer_3nm") %>% 
  dplyr::select(id_buff, geometry) %>% 
  rename(id = id_buff)

# Check the buffer:
mapview(pal_buff)

# Write as shapefile:
st_write(pal_buff, file.path(dir_git, 'spatial/shapefiles/pal_buffer_3nm.shp'), driver = "ESRI Shapefile")
```


# Join boundaries

Join shapefiles so we have one shapefile with all three boundaries  

```{r join-shoreline-monument}
# Double check the crs match
crs(marine_monument)
crs(shoreline)
crs(pal_buff)

# Bind shapefiles and calculate area in square kilometers
pal <- marine_monument %>%
  rbind(shoreline) %>% 
  rbind(pal_buff) %>% 
  mutate(area_m2 = st_area(.),
         area_km2 = area_m2/1000000) %>%
  separate(area_km2, into = c("area_km2", "units"), sep = " ") %>%
  dplyr::select(-area_m2, -units)

# View all three layers:
ggplot(pal) +
  geom_sf(aes(fill = id)) +
  theme_bw() +
  labs(fill = "OHI Palmyra Boundaries")

# View just the shoreline and buffer
pal %>% 
  filter(id != "marine_monument") %>% 
  ggplot(.) +
  geom_sf(aes(fill = id)) +
  theme_bw() +
  labs(fill = "OHI Palmyra Shoreline Buffer")

# Using mapview
mapview(pal, zcol="id")
```

# Check validity

```{r check-validity}
st_is_valid(pal) # all true 
```

# Save

```{r save}
st_write(pal, file.path(dir_git, 'spatial/shapefiles/pal_ohi_rgn.shp'), driver = "ESRI Shapefile")
```





