---
title: "create_rasters"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     # in_header: '~/github/ne-prep/src/templates/ohi_hdr.html' # update with own header
  pdf_document:
    toc: true
---

# Summary

Creating base rasters for data prep. Rasters were created at 100 m resolution and 1000 m resolution for both the 50nm boundary and the 3nm boundary.   

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(sf)
library(raster)
library(fasterize)
library(mapview)
library(utils)
library(here)

source(here('src/R/common.R'))

# for updating spatial 
pkgs <- c("sf", "raster", "fasterize")

# projections and crs
p4s_wgs84 <- '+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0'
p4s_nad83 <- '+proj=longlat +ellps=GRS80 +datum=NAD83 +no_defs +towgs84=0,0,0'
us_alb    <- raster::crs("+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs") 
```
# Methods 

## Extents for Cropping

Project shapefiles into both wgs and us albers and save the extent for both the 3nm buffer and the 50nm buffer in common.R.   
```{r extents, eval=F}
# Shapefiles
pal_boundary <- pal_monument
pal_buffer <- pal_buffer_3nm

pal_shp_sp <- pal_boundary %>%
  st_transform(crs = us_alb) %>% 
  as(.,"Spatial") # otherwise wont work in mask function

land <- pal_shoreline %>% 
  st_transform(crs = us_alb) %>% 
  as(., "Spatial") 

# Transform both shapefiles to wgs and us albers to save extents
pal_boundary_wgs <- pal_boundary %>% 
  st_transform(crs = p4s_wgs84)

pal_boundary_alb <- pal_boundary %>% 
  st_transform(crs = us_alb) %>% 
  mutate(rgn_id = 1)

pal_buffer_wgs <- pal_buffer %>% 
  st_transform(crs = p4s_wgs84)

pal_buffer_alb <- pal_buffer %>% 
  st_transform(crs = us_alb) %>% 
  mutate(rgn_id = 1)

# Extents for 50 nm boundary
wgs_ext_50nm <- raster::extent(pal_boundary_wgs)
alb_ext_50nm <- raster::extent(pal_boundary_alb) #this is for us_albers projection only

# Extents for 3 nm buffer
wgs_ext_3nm <- raster::extent(pal_buffer_wgs)
alb_ext_3nm <- raster::extent(pal_buffer_alb) #this is for us_albers projection only

# Check boundary 
plot(alb_ext_50nm)
plot(pal_boundary_alb, add = T)

# Check buffer
plot(alb_ext_3nm)
plot(pal_buffer_alb, add = T)
```


## 100 Meter Resolution  

### Base Rasters

Create a base raster clipped to the extent for the 3nm buffer
```{r create-base-rastser-3nm, eval=F}
# create a base raster with 100x100, crs = us albers and pal extent defined above
base_3nm <- raster(crs = us_alb, alb_ext_3nm, res = 100)
base_3nm[base_3nm] <- 1

plot(base_3nm)
plot(pal_buffer_alb, add = T)
```
Create a base raster clipped to the extent for the marine monument boundary 
```{r create-base-raster-50nm, eval=F}
# create a base raster with 100x100, crs = us_alb and pal extent defined above
base_50nm <- raster(crs = us_alb, alb_ext_50nm, res = 100)
base_50nm[base_50nm] <- 1

plot(base_50nm)
plot(pal_boundary_alb, add = T)
```

### Ocean Rasters

Mask out land to create ocean raster
```{r ocean-rasters-100m, eval=F}
# 3 nm
all_ras_3nm   <- mask(base_3nm, pal_buffer_alb)
ocean_ras_3nm <- mask(base_3nm, land, inverse = T)

plot(ocean_ras_3nm)
plot(land, add = T) 

# 50 nm
all_ras_50nm <- mask(base_50nm, pal_boundary_alb)
ocean_ras_50nm <- mask(base_50nm, land, inverse = T) 

plot(ocean_ras_50nm)
plot(land, add = T) 
```
### Save Rasters

Write out both ocean rasters for 50 nm boundary and the 3 nm buffer at 100 m resolution
```{r write_rasters-100m, eval=F}
writeRaster(ocean_ras_50nm, filename = "~/github/pal-prep/spatial/rasters/ocean_ras50nm_100m.tif", overwrite = T)
writeRaster(ocean_ras_3nm, filename = "~/github/pal-prep/spatial/rasters/ocean_ras3nm_100m.tif", overwrite = T)
```

## 1000 Meter Resolution

Repeat the above process for 1000 meter resolution   

```{r 1000m-rasters, eval=F}
# Create a base rasters with 1000x1000, crs = us_alb and Palmyra extent defined above
base_3nm_1000 <- raster(crs = us_alb, alb_ext_3nm, res = 1000)
base_3nm_1000[base_3nm_1000] <- 1

plot(base_3nm_1000)
plot(pal_buffer_alb)

base_50nm_1000 <- raster(crs = us_alb, alb_ext_50nm)
base_50nm_1000[base_50nm_1000] <- 1

plot(base_50nm_1000)
plot(pal_boundary_alb, add = T) 

# Ocean rasters
all_ras3_1k   <- mask(base_3nm_1000, pal_buffer_alb) 
ocean_ras3_1k <- mask(base_3nm_1000, land, inverse = T) 

all_ras50_1k   <- mask(base_50nm_1000, pal_boundary_alb)
ocean_ras50_1k <- mask(base_50nm_1000, land, inverse = T) 

plot(ocean_ras3_1k) 
plot(land, add=T) # Doesn't seem to be working?


plot(ocean_ras50_1k)
plot(land, add = T) # Doesn't seem to be working? 
```
```{r save-rasters-1000m, eval=F}
writeRaster(ocean_ras3_1k, filename = "~/github/pal-prep/spatial/rasters/ocean_ras3nm_1000m.tif", overwrite = T)
writeRaster(ocean_ras50_1k, filename = "~/github/pal-prep/spatial/rasters/ocean_ras50nm_1000m.tif", overwrite = T)
```
