---
title: "create_pal_region"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     # in_header: '~/github/ne-prep/src/templates/ohi_hdr.html' # update with own header
  pdf_document:
    toc: true
---

## Summary

This script shows how I created the shapefile for the Palmrya region. I used the following shapefiles:  

 * USFWS Special Area Designations - Marine National Monument Boundary
 * GADM country boundaries

## Data Sources

**Marine National Monument**  

* __Reference__: 
    * USFWS, NWRS, Division of Realty:. 2020. FWS Special Designation Areas - April 2020    <https://ecos.fws.gov/ServCat/Reference/Profile/116244>
        * __Accessed__: 5 June 2020   
* __Native data resolution__: Lands and waters administered by the USFWS in North America, U.S. Trust Territories and Possessions  
* __Format__:  Shapefile  

**Land Boundary**  


```{r setup, message = F, warning = F, results = 'hide'}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

source('~/github/pal-prep/src/R/common.R') ### Palmyra common.R

dir_git <- '~/github/pal-prep'
library(tidyverse)
library(sf)
library(mapview)
library(mapedit)
library(leaflet)
```


# Marine National Monument Boundary Shapefile Prep

The USFWS shapefile is for all of the U.S. and U.S. territories, we want to clip it to just include the National Marine Monument boundary and then save in a new folder  

```{r}
# Read in shapefile for all special areas designations
usfws_sp_area <- st_read(file.path(dir_git, 'spatial/raw_data_/usfws_special_areas'), layer = "SpecialDesignation")

# find only the pacific marine national monument palmyra unit 
marine_monument <- usfws_sp_area %>% 
  filter(CMPXNAME == "MARINE NATIONAL MONUMENTS OF THE PACIFIC") %>% 
  filter(COMMENTS == "KINGMAN REEF & PALMYRA ATOLL UNIT")

# View to make sure it's correct 
mapview(marine_monument)



```

