---
title: 'OHI+ Palmyra 2020: Research Goal Employment Layer'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    toc_depth: 2 
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     in_header: '../../../src/templates/ohi_hdr.html' 
  pdf_document:
    toc: true
---

# Summary

This script calculates the status of employment in conservation science on Palmyra Atoll.             

# Data 

Data for this layer was gathered through verbal communication with Alex Wegmann at The Nature Conservancy on July 8, 2020.         
# Setup

```{r setup, message = F, warning = F}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_goal   <- '~/github/pal-prep/prep/rs/v2020'
```


# Methods

## Create Raw Data

Palmyra supported 2 FTE positions in conservation science prior to 2019 and has supported 5 FTE positions in conservation science in 2019 and 2020.  

```{r create-raw-data, eval=F}

emp_raw <- data.frame(
  year = seq(2014,2020,1),
  positions = c(1,1,1,1,1,5,5)
)

# Save raw data 
write_csv(emp_raw, file.path(dir_goal, "raw/emp_raw.csv"))
```

## Calculate Current Status 

*Reference Point:* 5 full time equivalent (FTE) positions focused on conservation science at Palmyra Atoll      

```{r current-status, eval=F}

emp_raw <- read_csv(file.path(dir_goal, "raw/emp_raw.csv"))

# Define reference point 
ref_point <- 5

# Calculate the status 
emp_status <- emp_raw %>% 
  mutate(region_id = 1,
         status = positions / ref_point) %>% 
  dplyr::select(region_id, year, status)
  
```

## Save to Toolbox 

```{r save-to-toolbox, eval=F}
write_csv(emp_status, file.path(dir_scores, 'layers/rs_employment_status.csv'))
```

