---
title: 'Tourism and Recreation'
subtitle: 'OHI+ Palmyra 2020 Assessment'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3 
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     in_header: '~/github/pal-prep/src/templates/pal_hdr.html' 
  pdf_document:
    toc: true
---

# Summary 

This script status score for tourism and recreation based on the number of private visiting vessels per year from 2014 - 2020.    

# Data

Data for this layer is supposed to come from the boat visitation log on Palmyra Atoll. The log was not digitized at the time of the assessment, but estimates were obtained via email from Stefan Kropidlowski at the U.S. Fish and Wildlife Service on August 18, 2020.    

# Setup

```{r setup, message=F, warning=F, results='hide'}

library(tidyverse)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_goal   <- '~/github/pal-prep/prep/tr/v2020'
```


# Methods

The USFWS estimates about three private vessels visit Palmyra each year. We will use this estimate for all years from 2014 - 2020. These estimates should be updated once the boat visitation log on Palmyra Atoll is digitized.       

## Create Raw Data 

```{r create-raw-data, eval=F}
# Set years 
years <- 2014:2020

visits_raw <- data.frame(
  region_id = rep(1, length(years)), 
  year = years, 
  visits = rep(3, length(years))
)

# Save to raw 
write_csv(visits_raw, file.path(dir_goal, "raw/boat_visits_raw.csv"))
```

## Calculate Status 

**Reference Point:** Up to 6 private vessels visiting Palmyra Atoll per year. 

The current status will be scaled to the reference point using the following equation:      

$status~=~1~-~[(vessels_{y}~-vessels_{ref,y})~/~vessels_{ref,y}]$   

Where $vessels_{y}$ is the number of private vessels coming to the island in each year (y) and $vessel_{ref,y}$ is the reference point number of private vessels coming to the island in year (y). Any value greater than one will be given a status of one, and any value less than zero will be given a zero. This gives a perfect score to any year where the number of visiting vessels is at or below the reference point and a zero if the number of visiting vessels is double or more than double the reference point. This method of scoring is based on The Nature Conservancy's management goals to keep tourism at or below the reference point level and reflective of the view that having double the number of private vessels visiting each year would cause serious damage to the ecosystem.        

```{r calculate-status, eval=F}

vessels <- read_csv(file.path(dir_goal, "raw/boat_visits_raw.csv"))

tr_status <- vessels %>% 
  mutate(ref = 3,
         status = 1 - (visits - ref) / ref) %>% 
  mutate(status = ifelse(status > 1, 1, status)) %>% 
  dplyr::select(region_id, year, status)

# Save to output folder
write_csv(rao_status, file.path(dir_goal, "output/tr_status.csv"))
```

## Save to Toolbox

```{r save-to-toolbox, eval=F}
tr_status <- read_csv(file.path(dir_goal, "output/tr_status.csv"))
write_csv(tr_status, file.path(dir_scores, "layers/tr_status.csv"))
```

