---
title: 'Clean Waters: Soil Contamination Layer'
subtitle: 'OHI+ Palmyra 2020 Assessment'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3 
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     in_header: '~/github/pal-prep/src/templates/pal_hdr.html' 
  pdf_document:
    toc: true
---

# Summary

This script calculates a status score for soil contamination as part of the Clean Waters goal.  

# Data

Data for this layer comes from an elemental soil contamination study completed by U.S. Geological Survey at sites on Palmyra Atoll in 2008 and 2010. Data was published in 2017 and can be accessed [here](https://www.sciencebase.gov/catalog/item/5925b2cae4b0b7ff9fb3ca3d).  

Date Downloaded: July 28, 2020   

Citation: Struckhoff, M.A., Papoulias, D.M., Orazio, C.E., Annis, M.L., Shaver, D.K. and Tillitt, D.E., 2017, Geospatial data of elemental contamination at Palmyra Atoll, 2008 and 2010: U.S. Geological Survey data release, https://doi.org/10.5066/F74F1P00   

# Setup

```{r setup, message=F, warning=F, results='hide'}

library(tidyverse)
library(sf)
library(janitor)
library(ggplot2)
library(RColorBrewer)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_goal   <- '~/github/pal-prep/prep/cw/v2020'
```


# Methods

## Map Sample Sites 

Map the sample sites over the shapefile of the island.  

```{r map-sample-sites, message=F, warning=F, eval=F}

# Read in raw USGS shapefiles
usgs <- st_read(file.path(dir_anx, "_raw_data/USGS_data/XRF_samples_Palmyra_Atoll_2008_and_2010"),
                layer = "XRF_samples_Palmyra_Atoll_2008_and_2010") %>% 
  clean_names()

# Project shoreline file into WGS84
pal_wgs84 <- pal_shoreline %>% 
  st_transform(crs = wgs84)

# Map the study sites
ggplot() + 
  geom_sf(data = pal_wgs84, fill = "grey") + 
  geom_sf(data = usgs, color = "dodgerblue4")
```

## Reference Sites 

21 marine sediment samples and 7 soil samples were taken at sites believed to represent uncontaminated, natural concentrations of elements. These samples have a site_type of "reference". Save these locations as a separate data frame to compare against soil samples.   

```{r reference-sites, message=F, warning=F, eval=F}

ref_sites <- usgs %>% 
  filter(site_type == "Reference") %>% 
  mutate(soil_type = case_when(sample_matr == "Soil" ~ "soil",
                               site_sample == "Paradise 04" ~ "soil",
                               site_sample == "Strawn 02" ~ "soil", 
                               TRUE ~ "sediment")) %>% 
  pivot_longer(., cols = pb:p,
              names_to = "contaminant", values_to = "concentration") %>% 
  filter(contaminant != "fe" & contaminant != "ti" & contaminant != "p" & contaminant != "s") %>% 
  dplyr::select(site = site_sample, soil_type, year, contam = contaminant, 
                conc = concentration, geometry)

# Save to int folder 
st_write(ref_sites, file.path(dir_goal, "int/reference_sites.shp"), driver = 'ESRI Shapefile')
```

## Contaminants by Site

Assign a soil_type (marine or land) to the sample sites based on the reported soil or sediment classification. Most sites were sampled on a wet-weight basis, filter the 7 sites that were sampled on a dry-weight basis. Check if any of the sites have been sampled in both 2008 and 2010 and add a samp_repeated column to identify any sites with repeated sampling. Filter out the reference sites.     

```{r contamination-by-site, message=F, warning=F, eval=F}

# Read in raw USGS shapefiles
usgs <- st_read(file.path(dir_anx, "_raw_data/USGS_data/XRF_samples_Palmyra_Atoll_2008_and_2010"),
                layer = "XRF_samples_Palmyra_Atoll_2008_and_2010") %>% 
  clean_names()

# Color by site type to get an idea of what sites are offshore
ggplot() + 
  geom_sf(data = pal_wgs84, fill = "grey") + 
  geom_sf(data = usgs, aes(color = site_type))

# Wrangle data
sites <- usgs %>%
  mutate(dry = case_when(sample_matr == "Sludge-Dry" ~ "Yes",
                         sample_matr == "Sediment-Dry" ~ "Yes",
                         sample_matr == "Soil-Dry" ~ "Yes",
                         TRUE ~ "No")) %>% 
  filter(dry != "Yes") %>%  # Filter out the repeated sites with wet/dry samples
  filter(site_type != "Reference") %>%
  mutate(soil_type = case_when(sample_matr == "Soil" ~ "soil",
                               sample_matr == "Soil-Wet" ~ "soil",
                               sample_matr == "Sludge" ~ "soil",
                               sample_matr == "Sludge-Wet" ~ "soil",
                               sample_matr == "Solid" ~ "soil",
                               TRUE ~ "sediment")) # Everything else is sediment

# Check if we got all offshore sites correct
ggplot() + 
  geom_sf(data = pal_wgs84, fill = "grey") + 
  geom_sf(data = sites, aes(color = soil_type)) # Makes sense 


# Check if any sites have been sampled in both 2008 and 2009
check_years <- sites %>% 
  group_by(site_sample, year) %>% 
  tally()
# Looks like only Cooper 01, Cooper 02, Paradise 01, and Quail 09 are sampled in both years 

# Add repeated column and keep only necessary columns
usgs_sites <- sites %>% 
  mutate(samp_repeated = case_when(site_sample == "Cooper 01" ~ "Yes",
                                   site_sample == "Cooper 02" ~ "Yes",
                                   site_sample == "Paradise 01" ~ "Yes",
                                   site_sample == "Quail 09" ~ "Yes", 
                                   TRUE ~ "No"),
         area_km = pot_cont_are * 0.000001) %>% 
  dplyr::select(site = site_sample, soil_type, year, area_km, samp_repeated, 15:27)

# Site areas
site_area <- as.data.frame(usgs_sites) %>% 
  mutate(region_id = 1) %>% 
  dplyr::select(region_id, site, year, area_km) %>% 
  distinct() 


# Tidy format 
usgs_tidy <- pivot_longer(usgs_sites,
                          cols = pb:p,
                          names_to = "contaminant", values_to = "concentration") %>% 
  dplyr::select(site, soil_type, year, contam = contaminant, conc = concentration,
                repeated = samp_repeated, geometry)

# Check for the right number of observations - 13 contaminants total 
check <- usgs_tidy %>% 
  group_by(site) %>% 
  tally() # Only the four repeated sites should have 26, all others should be 13
## This looks good 

# Save shp and site areas to int folder 
st_write(usgs_tidy, file.path(dir_goal, "int/contaminants.shp"), driver = 'ESRI Shapefile')
write_csv(site_area, file.path(dir_goal, "int/site_areas.csv"))
```

## Reference Levels 

Two different reference levels will be used to scale the contamination scores from 0-1. A Threshold Effects Level (TEL) will be used to first scale values from 0 - 0.5. Any observed concentration equal to the TEL will receive a score of 0.5. The TEL are the lower end of values below which harmful impacts are not likely to be observed. The upper end of the scores will be based on the Probable Effect Level (PEL), this represents a value above which harmful impacts are likely to be observed. Any observed concentration above the PEL will be assigned a score of 1.  

### Assign TEL Reference Levels

TEL reference levels were assigned based on U.S. Environmental Protection Agency for marine sediments and the lowest established ecological soil screening level (Eco-SSL) for any taxonomic group. TEL and Eco-SSL values (mg/km dry weight) for the most frequently detected contaminants on Palmyra are listed in Table 1 of Struckhoff *et al* 2018.  

Citation: Struckhoff, M.A., Orazio, C.E., Tillitt, D.E., Shaver, D.K., and Papoulias, D.M. (2018). Mapping elemental contamination on Palmyra Atoll National Wildlife Refuge. *Marine Pollution Bulletin*, 128: 97-105. https://doi.org/10.1016/j.marpolbul.2017.12.065.  

Replicated from Table 1 of Struckhoff *e al* 2018. NA indicates a toxicity threshold has not been established.     

**Table 1**: Threshold effects levels and ecological soil screening levels for the nine contaminants found on Palmyra Atoll.   

| Contaminant | Symbol | TEL   | Eco-SSL |
|-------------|--------|-------|---------|
| Antimony    | Sb     | NA    | 78      |
| Arsenic     | As     | 7.24  | 18      |
| Chromium    | Cr     | 52.3  | 26      |
| Colbalt     | Co     | NA    | 13      |
| Copper      | Cu     | 118.7 | 28      |
| Lead        | Pb     | 30.24 | 11      |
| Manganese   | Mn     | NA    | 220     |
| Nickel      | Ni     | 15.9  | 38      |
| Zinc        | Zn     | 124   | 46      |

A total of 13 contaminants were sampled during the study. Contaminants not included in the table above are iron (Fe), titanium (Ti), sulfur (S), and phosphorus (P). Iron is generally considered non-toxic and titanium, sulfur, and phosphorus were detected but do not have established TEL or Eco-SSL levels, therefore they will excluded from the analysis.  

The table reports thresholds in mg/kg of dry weight, however most of the samples collected were made on moist samples. Struckoff *et al* 2018 notes:   

"Soil moisture values between 15 and 25% can reduce measured sample concentrations from 70 to 80% relative to lab-confirmed concentrations (U.S. Environmental Protection Agency, 2015). Given that most measurements were made on moist samples, XRF values reported above TELs or SSLs are conservative estimates of threshold exceedance."  

```{r tel-ref-levels, message=F, warning=F, eval=F}

contaminant <- st_read(file.path(dir_goal, "int"),
                layer = "contaminants") %>% 
  filter(contam != "fe" & contam != "ti" & contam != "p" & contam != "s")

# Set reference levels for each contaminant
# Where TEL is unknown, the lowest eco-ssl is used in its place 
ref_tel <- data.frame(
  contam = rep(c("sb", "as", "cr", "co", "cu", "pb", "mn", "ni", "zn"), each = 2),
  soil_type = rep(c("sediment", "soil"), 9),
  lower_type = rep(c("tel", "eco-ssl"), 9),
  lower_limit = c(78, 78, 7.24, 18, 52.3, 26, 13, 13, 118.7, 28,
                30.24, 11, 220, 220, 15.9, 38, 124, 46)
)

# Add reference levels to contaminants data
contam_tels <- contaminant %>% 
  left_join(ref_tel, by = c("contam","soil_type")) # Soil uses eco-ssl, marine sediments use tel
```

### Assign PEL Reference Levels 

For soil samples, the Canadian freshwater contamination PELs (mg/kg dry weight) are used, as these are more conservative than the marine contamination levels. Consensus-based freshwater PEL levels were established in 2000. Concentrations were obtained from Table 3 in MacDonald *et al* 2000.   

Citation: MacDonald, D.D., Ingersol, C.G., and Berger, T.A. (2000). Development and Evaluation of Consensus-Based Sediment Quality Guidelines for Freshwater Contamination. *Archives of Environmental Contamination and Toxicology.* 39:20-31. DOI: 10.1007/s002440010075   

**Table 2**: Freshwater probable effect levels for contaminants found on Palmyra Atoll.   

| Contaminant | Symbol | PEL   | 
|-------------|--------|-------|
| Arsenic     | As     | 17    |
| Chromium    | Cr     | 90    |
| Copper      | Cu     | 197   |
| Lead        | Pb     | 91.3  |
| Nickel      | Ni     | 36    |
| Zinc        | Zn     | 315   |

Marine sediment PEL levels (mg/kg of dry weight) are obtained from NOAA guidelines established in 1999. These are called Effect Range Median (ERM) levels, which is the NOAA equivalent of PELs. Concentrations were obtained from Table 1.   

Citation: National Oceanic and Atmospheric Administration. (1999). Sediment Quality Guidelines developed for the National Status and Trend Program. 

**Table 3**: Marine sediment probable effect levels for contaminants found on Palmyra Atoll.   

| Contaminant | Symbol | PEL   | 
|-------------|--------|-------|
| Arsenic     | As     | 70    |
| Chromium    | Cr     | 370   |
| Copper      | Cu     | 270   |
| Lead        | Pb     | 218   |
| Nickel      | Ni     | 51.6  |
| Zinc        | Zn     | 410   |

No PELs are available for Antimony, Colbalt, and Manganese. For these we will approximate an upper limit that is one order of magntiude higher than the established TEL or Eco-SSL.  

```{r pel-ref-levels, message=F, warning=F, eval=F}

# Set reference levels for each contaminant
# Where PEL is not established use one order of magnitude higher than TEL
ref_pel <- data.frame(
  contam = rep(c("sb", "as", "cr", "co", "cu", "pb", "mn", "ni", "zn"), each = 2),
  soil_type = rep(c("soil", "sediment"), 9),
  upper_type = rep(c("pel", "erm"), 9),
  upper_limit = c(780, 780, 17, 70, 90, 370, 130, 130, 197, 270,
                91.3, 218, 2200, 2200, 36, 51.6, 315, 410)
)

# Add PEL reference levels to TEL ref levels
contam_ref <- contam_tels %>% 
  left_join(ref_pel, by = c("contam","soil_type")) # Soil use pel, marine sediments use erm

# Save to int folder
st_write(contam_ref, file.path(dir_goal, "int/contaminant_refs.shp"), driver = 'ESRI Shapefile')
```


## Calculate Status

**Reference Point:** The TELs and PELs for each individual contaminant are used as the reference point values.   

The current status for soil contamination will be linearly scaled 0-1 using the TEL and PEL reference levels as midpoint (0.5) and end point (1) values, respectively. Values above 1 will be capped at 1.  

A status is then found for each site using a geometric mean of the status of all contaminants at that site. The overall status score is found for each year as an area weighted mean. This will be used for the soil contamination pressure layer. The CW status score will be the inverse, found by 1 - status.   

```{r calc-status, eval=F}

contam_refs <- st_read(file.path(dir_goal, "int"),
                layer = "contaminant_refs")

site_areas <- read_csv(file.path(dir_goal, "int/site_areas.csv"))

# Find status score for each contaminant at each site 
contam_status <- contam_refs %>% 
  mutate(c_status = ifelse(conc > lwr_lmt, 0.5, (conc / lwr_lmt) / 2), 
         c_status = ifelse(c_status == 0.5, 0.5 + (conc / uppr_lm), c_status), 
         c_status = ifelse(c_status > 1, 1, c_status), # Cap values above 1 at 1
         c_status = ifelse(c_status == 0, 0.001, c_status), # Prevent mean from going to 0
         region_id = 1) %>% 
  dplyr::select(region_id, site, year, contam, c_status) 

# Define function for geometric mean 
geometric.mean <- function (x, na.rm = TRUE) {
    if (is.null(nrow(x))) {
      exp(mean(log(x), na.rm = TRUE))
    }
    else {
      exp(apply(log(x), 2, mean, na.rm = na.rm))
    }
}

# Find geometric mean of contaminant status at each site  
site_status <- contam_status %>% 
  group_by(site, year) %>% 
  summarize(s_status = geometric.mean(c_status))

# Find overall status for each year as the area weighted mean of each site
status <- site_status %>% 
  group_by(site, year) %>% 
  left_join(site_areas, by = c("site", "year")) %>% 
  ungroup() %>% 
  group_by(year) %>% 
  summarize(value = weighted.mean(s_status, area_km)) %>% 
  mutate(region_id = 1,
         status = 1 - value) %>% 
  dplyr::select(region_id, year, status)

# The 2008 and 2010 statuses are similar, we will use 2010 as an estimate for every year 2014-2020
sc_status <- data.frame(
  region_id = 1,
  year = 2010, 
  status = status$status[status$year == 2010]
)

# Save to output folder
write_csv(sc_status, file.path(dir_goal, "output/sc_status.csv"))
```


## Save to Toolbox 

```{r save-to-toolbox, message=F, warning=F, eval=F}
sc_status <- read_csv(file.path(dir_goal, "output/sc_status.csv"))
write_csv(sc_status, file.path(dir_scores, "layers/cw_contamination_status.csv"))
```

# Pressure Layer

## Calculate Pressure Layer

Soil contamination will also be a pressure layer where higher scores indicate more contamination. This is the inverse of the status score calculated above.  

```{r debris-pressure-layer, message=F, warning=F, eval=F}
sc_status <- read_csv(file.path(dir_goal, "output/sc_status.csv"))

# Use same pressure score from 2014-2020
years  <- 2014:2020
status <- sc_status$status

# Find pressure score - inverse of status score 
prs <- data.frame(
  region_id = rep(1, length(years)), 
  year = years,
  pressure_score = rep(1 - status, length(years))
)

# Save to output
write_csv(prs, file.path(dir_goal, "output/po_soil.csv"))
```

## Save Pressure to Toolbox 

```{r save-prs-toolbox, message=F, warning=F, eval=F}
po_soil <- read_csv(file.path(dir_goal, "output/po_soil.csv"))
write_csv(po_soil, file.path(dir_scores, "layers/prs_po_contamination.csv"))
```


