---
title: 'OHI+ Palmyra 2020: Management Pressure'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    toc_depth: 2 
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     in_header: '../../../src/templates/ohi_hdr.html' 
  pdf_document:
    toc: true
---

# Summary 

This script calculates the pressures from a lack of management on Palmyra Atoll. This layer is the inverse of the management resilience layer. 

# Data 

Data to inform this layer was obtained from verbal communications with Alex Wegmann at The Nature Conservancy on July 31, 2020. 

# Setup 

```{r setup, message = F, warning = F}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_prs   <- '~/github/pal-prep/prep/pressures/management/v2020'
dir_mgmt  <- '~/github/pal-prep/prep/resilience/ecological/habitat_destruction/v2020'
```

# Methods 

## Pressure Score 

This layer is the inverse of the habitat destruction management resilience layer. This layer includes management scores for both inland and offshore areas. Information on how the score for the management resilience layer was derived can be found in the regulatory resiliences script.   

```{r pressure-score, eval=F}
# Read in management score layers 
mgmt_res <- read_csv(file.path(dir_mgmt, "output/hd_mgmt.csv")) 

# Find the pressure score 
mgmt_pressure <- mgmt_res %>% 
  mutate(pressure_score = 1 - resilience_score) %>% 
  dplyr::select(-resilience_score)

# Save to output folder
write_csv(mgmt_pressure, file.path(dir_prs, "output/mgmt_pressure.csv"))
```

## Save to Toolbox 

```{r save-to-toolbox, eval=F}
mgmt_pressure <- read_csv(file.path(dir_prs, "output/mgmt_pressure.csv"))
write_csv(mgmt_pressure, file.path(dir_scores, "layers/prs_mgmt.csv"))
```

