---
title: 'OHI+ Palmyra 2020: Fishing Regulations Resilience Layer'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    toc_depth: 3 
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     in_header: '../../../src/templates/ohi_hdr.html' 
  pdf_document:
    toc: true
---

# Summary 

Regulatory resilience is meant to describes the institutional measures (e.g., rules, regulations, and laws) designed to address ecological pressures. This script calculates resilience from regulations on fishing activities on Palmyra Atoll. Primary regulations on fishing actvities come from the designation of marine areas as protected, which prohibits fishing within these areas.     

# Data 

Data to inform the preparation of these layers was obtained through verbal communications with Alex Wegmann of The Nature Conservancy on July 31, 2020.  

# Setup 

```{r setup, message = F, warning = F}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_res   <- '~/github/pal-prep/prep/resilience/ecological/fishing_pressure/v2020'
dir_lsp   <- '~/github/pal-prep/prep/sp/lsp/v2020'
```


# Methods 

## Protected Areas

One regulatory mechanism to boost resilience to fishing pressures is designation of protected areas. The percent of protected areas is calculated as part of the Lasting Special Places sub-goal of the Sense of Place goal. For fishing pressure, only marine protected areas are included.  

```{r protected-areas, eval=F}
# Get offshore and inland LSP status scores 
offshore <- read_csv(file.path(dir_lsp, "output/offshore_status.csv")) %>% 
  mutate(dimension = "offshore")

# Use the offshore score for fishing pressure (only need years 2014-2020)
fp_mpa <- offshore %>% 
  filter(year >= 2014) %>% 
  dplyr::select(region_id, year, resilience_score = status)

# Save files to output folder
write_csv(fp_mpa, file.path(dir_res, "output/fp_reg.csv"))
```

## Save to Toolbox

```{r save-to-toolbox, eval=F}
fp_reg <- read_csv(file.path(dir_res, "output/fp_reg.csv"))
write_csv(fp_reg, file.path(dir_scores, "layers/res_fp_reg.csv"))
```

