---
title: 'OHI+ Palmyra 2020: Ecosystem Resilience'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    toc_depth: 3 
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     in_header: '../../../src/templates/ohi_hdr.html' 
  pdf_document:
    toc: true
---

# Summary 

Ecosystem resilience is meant to address the integrity of ecosystems. Ecosystem integrity, e.g., food web integrity, is estimating the relative condition of assessed species. Ecosystems that are more intact and support higher biodiversity are more likely to be resilient to ecological pressures. This script calculates ecosystem resilience using the Species sub-goal score. 

# Data 

Species biodiversity is calculated using the 50 nm Pacific Marine National Monument boundary. Methods and data used to obtain the species sub-goal score can be found in data prep scripts in `pal-prep/prep/bd/spp` or in the Methods document.    

# Setup 

```{r setup, message = F, warning = F}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_res   <- '~/github/pal-prep/prep/resilience/ecological/ecosystem/v2020'
dir_spp   <- '~/github/pal-prep/prep/bd/spp/v2020'
```


# Methods 

## Resilience Score 

The resilience score will use the same score as species status, calculated in the `3_spp_status.Rmd` script.  

```{r eco-res-score, eval=F}
# Use the same status score as the toolbox (for now)
spp_status <- read_csv(file.path(dir_spp, "output/spp_status.csv")) 

# Create new resilience layer 
eco_spp <- spp_status %>% 
  rename(resilience_score = status)

# Save to output folder
write_csv(eco_spp, file.path(dir_res, "output/eco_spp_diversity.csv"))
```

## Save to Toolbox 

```{r save-to-toolbox, eval=F}
eco_spp <- read_csv(file.path(dir_res, "output/eco_spp_diversity.csv"))
write_csv(eco_spp, file.path(dir_scores, "layers/res_eco_spp_diversity.csv"))
```

