---
title: 'OHI+ Palmyra 2020 - Lasting Special Places subgoal'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    toc_depth: 3 
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     in_header: '../../../src/templates/ohi_hdr.html' 
  pdf_document:
    toc: true
---

# Summary

This script calculates the current status of lasting special places on Palmyra Atoll. While lasting special places are challenging to quantify, this goal assumes areas that are protected represent special places (i.e. the effort to protect them suggests they are important places) (Halpern *et al* 2012).    

# Data 

Data for this goal was given verbally by Alex Wegmann at The Nature Conservancy on June 29, 2020. Palmyra Atoll is 100% protected, the Wildlife Refuge at Palmyra was [established in 2001](https://www.fisheries.noaa.gov/pacific-islands/habitat-conservation/pacific-remote-islands-marine-national-monument). 

# Setup

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(sf)
library(here)

# Set filepaths and source functions
source(here('src/R/common.R'))
source(ohicore)

dir_goal   <- '~/github/pal-prep/sp/lsp/v2020'
```

# Methods

## Offshore Region Area  

Create the offshore region area output file based on the Palmyra shapefiles. The offshore region is based on a 3 nm buffer around the island.  

```{r create-rgn-area-offshore, eval=F}

# Calculate region area from the 3nm buffer shapefile:
buff_3nm <- pal_buffer_3nm %>% 
  mutate(area_m2 = st_area(.),
         area_km2 = area_m2/1000000) %>%
  separate(area_km2, into = c("area_km2", "units"), sep = " ") %>%
  dplyr::select(-area_m2, -units)

area_3nm <- as.numeric(buff_3nm$area_km2)

# Create first output file
rgn_area_offshore3nm <- data.frame(rgn_id = 1, area = area_3nm)

# Save to output folder:
write_csv(rgn_area_offshore3nm, file.path(dir_goal, 'output/rgn_area_offshore3nm'))
```

## Inland Region Area

Create inland region area output file. The entire land area of Palmyra Atoll will be included in the inland region.  

```{r create-rgn-area-inland, eval=F}

# Calculate region area from the land shapefile:
land <- pal_shoreline %>% 
  mutate(area_m2 = st_area(.),
         area_km2 = area_m2/1000000) %>%
  separate(area_km2, into = c("area_km2", "units"), sep = " ") %>%
  dplyr::select(-area_m2, -units)

# whole island is slightly over 2km so we can use the whole area 
area_1km <- as.numeric(land$area_km2)

# Create the output file
rgn_area_inland1km <- data.frame(rgn_id = 1, area = area_1km)

# Save to output folder:
write_csv(rgn_area_inland1km, file.path(dir_goal, 'output/rgn_area_inland1km'))
```

## Offshore Protected Areas 

Find the total offshore area that is protected on Palmyra. Since the establishment of the Wildlife Refuge in 2001, the entire island offshore area has been protected.   

```{r create-lsp-prot-area-offshore, eval=F}
# Wildlife Refuge was established in 2001 so I'll use years 2001-2019
years <- seq(2001,2020,1)

# Create output file
lsp_prot_area_offshore3nm <- data.frame(rgn_id = rep(1, length(years)),
                                        year = years,
                                        a_prot_3nm = rep(area_3nm, length(years)))

# Save to output folder
write_csv(lsp_prot_area_offshore3nm, file.path(dir_github, 'prep/sp/lsp/v2020/output/lsp_prot_area_offshore3nm'))
```

## Inland Protected Areas 

Find the total inland area that is protected on Palmyra. The Nature Conservancy bought Palmyra in 2000 and the entire island area has been protected.  

```{r create-lsp-prot-area-inland, eval=F}
# Use same years as offshore
years <- seq(2001,2020,1)

# Create output file
lsp_prot_area_inland1km <- data.frame(rgn_id = rep(1, length(years)),
                                        year = years,
                                        a_prot_1km = rep(area_1km, length(years)))

# Save to output folder
write_csv(lsp_prot_area_inland1km, file.path(dir_github, 'prep/sp/lsp/v2020/output/lsp_prot_area_inland1km'))
```

## Calculate Status 

*Reference Point:* 99% of Palmyra Atoll is protected   

The lasting special places status is calculated for offshore + inland areas relative to the reference point, and the average status of both is used as the overall status score.  

$status_{offshore}~=~pct_{offshore}~/~pct~ref_{offshore}$   

$status_{inland}~=~pct_{inland}~/~pct~ref_{inland}$    

Where $pct_{offshore}$, $pct_{inland}$ is the percent of offshore and inland areas, respectively, that are protected and $pct~ref_{offshore}$, $pct~ref_{inland}$ are the reference point percentages of protected areas offshore and inland, respectively.  

Finally, the status of offshore and inland areas are averaged to find the overall status.  

$status~=~(status_{offshore}~+status_{inland})~/~2$

```{r calc_status, eval=F}
# Read in the output data files:  
rgn_area_inland1km <- read_csv(file.path(dir_github, 'prep/sp/lsp/v2020/output/rgn_area_inland1km')) %>% 
  mutate(layer = "rgn_area_inland1km")

rgn_area_offshore3nm <- read_csv(file.path(dir_github, 'prep/sp/lsp/v2020/output/rgn_area_offshore3nm')) %>% 
  mutate(layer = "rgn_area_offshore3nm")

lsp_prot_area_inland1km <- read_csv(file.path(dir_github, 'prep/sp/lsp/v2020/output/lsp_prot_area_inland1km'))

lsp_prot_area_offshore3nm <- read_csv(file.path(dir_github, 'prep/sp/lsp/v2020/output/lsp_prot_area_offshore3nm'))

# Set scenario year 
scenario_year = 2020

# Set reference points - 99% of all Palmyra area protected 
ref_pct_mpa <- 99 # marine protected area
ref_pct_cpa <- 99 # coastal protected area

# calculate the percent of offshore protected areas and the status
offshore <- lsp_prot_area_offshore3nm %>% 
  mutate(area_offshore3nm = rgn_area_offshore3nm$area,
         scenario_year = 2020) %>% 
  dplyr::rename(mpa = a_prot_3nm) %>% 
  mutate(pct_mpa  = pmin(mpa  / area_offshore3nm * 100, 100),
         status   = pmin(pct_mpa / ref_pct_mpa, 1)) %>% 
  dplyr::select(region_id = rgn_id,
                year,
                status)

# calculate the percent of inland protected areas and the status
inland <- lsp_prot_area_inland1km %>% 
  mutate(area_inland1km = rgn_area_inland1km$area,
         scenario_year = 2020) %>% 
  dplyr::rename(cpa = a_prot_1km) %>% 
  mutate(pct_cpa  = pmin(cpa  / area_inland1km * 100, 100),
         status   = pmin(pct_cpa / ref_pct_cpa, 1)) %>% 
  dplyr::select(region_id = rgn_id,
                year,
                status)
```

## Save To Toolbox 

Save the two output to the toolbox as `lsp_offshore.csv` and `lsp_inland.csv`      

```{r save-to-toolbox, eval=F}
write_csv(offshore, file.path(dir_scores, 'layers/lsp_offshore.csv'))
write_csv(inland, file.path(dir_scores, 'layers/lsp_inland.csv'))
```

