---
title: 'Coastal Protection Trend'
subtitle: 'OHI+ Palmyra 2020 Assessment'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3 
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     in_header: '~/github/pal-prep/src/templates/pal_hdr.html' 
  pdf_document:
    toc: true
---

# Summary 

This script calculates the trend in coastal protection based on the trend in the condition of coral reef and rainforest habitats.        

# Data 

The coastal protection trend will use the same data as the coral condition trend and the rainforest condition trend calculated for the Biodiversity Habitats sub-goal. Refer to the [coral trend data prep](https://raw.githack.com/OHI-4site/pal-prep/gh-pages/prep/bd/hab/v2020/coral_trend.html) and the [rainforest trend data prep](https://raw.githack.com/OHI-4site/pal-prep/gh-pages/prep/bd/hab/v2020/rainforest_trend.html) for additional information.   

# Setup 

```{r setup, message=F, warning=F, results='hide'}

library(tidyverse)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_goal   <- '~/github/pal-prep/prep/hs/cp/v2020'
dir_hab    <- '~/github/pal-prep/prep/bd/hab/v2020'
```

# Methods 

## Coastal Protection Trend 

The coastal protection trend are based on the coral condition and rainforest condition trends, which were calculated for the Biodiversity Habitats sub-goal in `coral_trend.Rmd` and `rainforest_trend.Rmd`. The trend will be calculated using the same method as the coastal protection status.   

```{r cp-trend, eval=F}

# Read in trend data
coral_trend      <- read_csv(file.path(dir_hab, "output/coral_trend.csv"))
rainforest_trend <- read_csv(file.path(dir_hab, "output/rainforest_trend.csv")) 

# Read in cp data 
cp_data <- read_csv(file.path(dir_goal, "int/cp_data.csv"))

# Combine trend data 
trend <- coral_trend %>% 
  rbind(rainforest_trend) 

# Add trend data to cp_data 
cp_data_trend <- cp_data %>% 
  dplyr::full_join(trend, by = c("region_id", "habitat")) %>% 
  dplyr::select(-year)

# Calculate trend:
cp_trend <- cp_data_trend %>% 
  group_by(region_id) %>%
  dplyr::summarize(
          trend = sum(rank * trend * extent, na.rm = TRUE) /
                  (sum(extent * rank, na.rm = TRUE))) %>% 
  mutate(year = 2020) %>% 
  dplyr::select(region_id, year, trend)

# Save to output
write_csv(cp_trend, file.path(dir_goal, "output/cp_trend.csv"))
```

## Save to Toolbox 
```{r save-to-toolbox, eval=F}
cp_trend <- read_csv(file.path(dir_goal, "output/cp_trend.csv"))
write_csv(cp_trend, file.path(dir_scores, "layers/hs_coastal_protection_trend.csv"))
```


