---
title: 'Coastal Protection Sub-goal'
subtitle: 'OHI+ Palmyra 2020 Assessment'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3 
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
     in_header: '~/github/pal-prep/src/templates/pal_hdr.html' 
  pdf_document:
    toc: true
---

# Summary 

This script calculates the status of coastal protection from coral reef habitats on Palmyra Atoll.  

# Data 

Coastal protection status is found using coral reef extent and condition and rainforest extent and condition.    

*Coral Reef Extent and Condition*   

Coral reef extent and condition data are the same data used in the coral habitat status layer. Coral cover maps were downloaded from the National Centers for Coastal Ocean Science (NCCOS) benthic habitat mapping project for Palmyra Atoll, found [here](https://products.coastalscience.noaa.gov/collections/benthic/e58palmyra/#horizontalTab3)   

Habitat maps were completed in December 2011 and data were downloaded on June 30, 2020.  

Citation: U.S. Department of Commerce, National Oceanic and Atmospheric Administration, National Center for Coastal Ocean Sciences. (2011). Palmyra Habitat Maps [Data file]. Retrieved from https://products.coastalscience.noaa.gov/collections/benthic/e58palmyra/
<br>    
*Rainforest Extent and Condition*

The most recent estimates for rainforest coverage on Palmyra Atoll comes from a 2019 study quantifying coconut palm extent.  

Citation: Burnett, M.W. *et al.* (2019). "Quantifying coconut palm extent on Pacific islands using spectral textural analysis of very high resolution imagery". *International Journal of Remote Sensing* 40(19): 7329-7355   
  [Available here](https://www.tandfonline.com/doi/epub/10.1080/01431161.2019.1594440?needAccess=true). 
  
The data come from Table A1 downloaded on July 8, 2020 from [here](https://www.tandfonline.com/doi/figure/10.1080/01431161.2019.1594440?scroll=top&needAccess=true).

# Setup 

```{r setup, message=F, warning=F, results='hide'}

library(tidyverse)
library(here)

# Source and set file paths
source(here('src/R/common.R'))

dir_goal   <- '~/github/pal-prep/prep/hs/cp/v2020' 
dir_hab    <- '~/github/pal-prep/prep/bd/hab/v2020'
```


# Methods 

## Condition

Condition is the current condition of coral reef habitats and rainforest habitats relative to the reference point.   

$C_{condition}~=~C_{current}~/~C_{ref}$   

Where $C_{current}$ is the current status of the habitat and $C_{ref}$ is the reference point. This is the same as the coral status and the rainforest status calculated in the Biodiversity Habitats sub-goal.   

```{r condition, eval=F}

# Habitat status layers
coral_status      <- read_csv(file.path(dir_hab, "output/coral_status.csv"))
rainforest_status <- read_csv(file.path(dir_hab, "output/rainforest_status.csv"))

# Find a single estimate of condition by averaging the status of each zone 
coral_condition <- coral_status %>% 
  group_by(region_id) %>% 
  summarize(status = weighted.mean(status, area_km)) %>% 
  mutate(habitat = "coral") %>% 
  dplyr::select(region_id, habitat, condition = status)

# Rainforest status 
rainforest_condition <- rainforest_status %>% 
  dplyr::select(region_id, habitat, condition = status)

# Combine condition layers 
condition <- coral_condition %>% 
  rbind(rainforest_condition)

# Save to int 
write_csv(condition, file.path(dir_goal, "int/cp_condition.csv"))
```

## Extent 

Use the extent calculated as parts of the Biodiversity: Habitats sub-goal.    

```{r extent, eval=F}

# Habitat extents
coral_extent      <- read_csv(file.path(dir_hab, "output/coral_extent.csv"))
rainforest_extent <- read_csv(file.path(dir_hab, "output/rainforest_extent.csv")) %>% 
  filter(year == 2019)

# Combine extent layers
extent <- coral_extent %>% 
  rbind(rainforest_extent) %>% 
  dplyr::select(region_id, habitat, extent = km2) 

# Save to int folder
write_csv(extent, file.path(dir_goal, "int/cp_extent.csv"))
```

## Add Protectiveness Ranks 

Combine the condition and extent layers and add protectiveness ranks for use in calculating the status.   

Protectiveness ranks are based on those used in the [OHI Global assessment](http://htmlpreview.github.io/?https://github.com/OHI-Science/ohi-global/published/documents/methods/Supplement.html#63_coastal_protection). Although the OHI global does not include native atoll forest as a category, we ranked it equal to mangroves or salt marshes given they occupy similar shoreline regions.   

```{r protectiveness-ranks, eval=F}

# Read in the combined condition and extent files
condition <- read_csv(file.path(dir_goal, "int/cp_condition.csv"))
extent    <- read_csv(file.path(dir_goal, "int/cp_extent.csv"))

# Define protectiveness ranks
ranks <- c('coral' = 3, 'rainforest' = 4)

# Combine data 
cp_data <-  extent %>%
  dplyr::full_join(condition, by = c("region_id", "habitat")) %>% 
  mutate(rank = ranks[habitat])

# Save to int folder
write_csv(cp_data, file.path(dir_goal, "int/cp_data.csv"))
```


## Calculate Status

**Reference Point:** The coral and rainforest condition will use the same reference points as the Habitats sub-goal.    

The coastal protection status is calculated as the relative health of the habitats that provide coastal protection weighted by their area and protectiveness rank. Coral reefs are the only habitat included in the coastal protection goal for Palmyra, and the status is found by the following equation:  

$status = \frac { \displaystyle\sum _{ k=1 }^{ N }{ { (h }_{ k } } \times { w }_{ k }\times { A }_{ k }) }{ \displaystyle\sum _{ k=1 }^{ N }{ { (w }_{ k }\times { A }_{ k }) }  }$   

where $w$ is the rank weight of the habitat’s protective ability, $A$ is the area within a region for each $k$ habitat type, and $h$ is a measure of each habitat's condition.   

```{r calculate-status, eval=F}
 
# Read in cp data
cp_data <- read_csv(file.path(dir_goal, "int/cp_data.csv"))

# Calculate status 
cp_status <- cp_data %>% 
  group_by(region_id) %>%
  summarize(status = pmin(1, sum(rank * condition * extent, na.rm = TRUE) /
                            (sum(extent * rank, na.rm = TRUE)))) %>% 
  mutate(year = 2020) %>% 
  dplyr::select(region_id, year, status)


# Save to output
write_csv(cp_status, file.path(dir_goal, "output/cp_status.csv"))
```

## Save to Toolbox 

```{r save-to-toolbox, eval=F}
cp_status <- read_csv(file.path(dir_goal, "output/cp_status.csv"))
write_csv(cp_status, file.path(dir_scores, 'layers/hs_coastal_protection_status.csv'))
```

