---
title: "corals"
author: "Gage Clawson"
date: "7/9/2020"
output: html_document
---

# Summary

This script generates the extent of coral for Palmyra.

***
## Data Source 

**Downloaded**: 07/25/2019

**Description**:  
Global Distribution of Coral Reefs
https://data.unep-wcmc.org/datasets/1
Reported at spatial cell scale. 

This dataset shows the global distribution of coral reefs in tropical and subtropical regions. It is the most comprehensive global dataset of warm-water coral reefs to date, acting as a foundation baseline map for future, more detailed, work. This dataset was compiled from a number of sources by UNEP World Conservation Monitoring Centre (UNEP-WCMC) and the WorldFish Centre, in collaboration with WRI (World Resources Institute) and TNC (The Nature Conservancy). Data sources include the Millennium Coral Reef Mapping Project (IMaRS-USF and IRD 2005, IMaRS-USF 2005) and the World Atlas of Coral Reefs (Spalding et al. 2001).

**Time range**: 1954-2018


***
# Methods
Reclassify the coral extent data into a mask of 1 or NA, and then compute zonal statistics for the count of cells within Palmyra that have coral and then convert into km2.


## Setup
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(readr)      # for read_csv()
library(raster)
library(here)
library(sf)
library(fasterize)
library(tidyverse)

source(file.path('~/github/ohiprep_v2020/workflow/R/common.R'))

goal     <- 'globalprep/hab_coral/v2019'
dir_git  <- file.path('~/github/ohiprep_v2019', goal)
dir_wcmc <- file.path(file.path(dir_M, 'git-annex/globalprep/_raw_data/wcmc_coral'))
ohi_rasters() # call the region zones raster
```

```{r, echo = FALSE, eval = FALSE}

v4_coral_pts <- sf::st_read(dsn = file.path(dir_wcmc, "14_001_WCMC008_CoralReefs2018_v4/01_Data"), layer = "WCMC008_CoralReef2018_Pt_v4")

v4_coral_py <- sf::st_read(dsn = file.path(dir_wcmc, "14_001_WCMC008_CoralReefs2018_v4/01_Data"), layer = "WCMC008_CoralReef2018_Py_v4")

v4_coral_metadata <- sf::st_read(dsn = file.path(dir_wcmc, "14_001_WCMC008_CoralReefs2018_v4/01_Data"), layer = "Metadata_CoralReefs")

```

**Convert coral shapefiles into same CRS as our region zones raster**
```{r, echo = FALSE, eval = FALSE}

moll_crs <- crs(zones, asText = TRUE)

v4_coral_pts_moll <- st_transform(v4_coral_pts, crs = moll_crs) #project points shapefile to match zones crs


crs(v4_coral_pts_moll) #check to see it worked

v4_coral_py_moll <- st_transform(v4_coral_py, crs = moll_crs)

## Subset global dataset to match Palmyra shape. 
source(here('src/R/common.R'))

us_alb    <- raster::crs("+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs") 

pal_buffer <- pal_buffer_3nm


global_py_alb <- v4_coral_py_moll %>%
  st_transform(crs = us_alb)

pal_buffer_alb <- pal_buffer %>% 
  st_transform(crs = us_alb) %>% 
  dplyr::mutate(rgn_id = 1)

alb_ext_3nm <- raster::extent(pal_buffer_alb) #this is for us_albers projection only


pal_py <- st_crop(global_py_alb, alb_ext_3nm)

plot(pal_py[21])

## this data came from the Millennium Coral Reef Mapping Project.. ending in 2002 according to the metadata: http://imars.marine.usf.edu/MC/


global_pt_alb <- v4_coral_pts_moll %>%
  st_transform(crs = us_alb)

pal_pt <- st_crop(global_pt_alb, alb_ext_3nm) ## nothing in the points data. Will only work with polygons then...

pal_3nm_rast <- raster::raster("../../../../spatial/rasters/ocean_ras_3nm.tif")
plot(pal_3nm_rast)


```

**Fasterize/rasterize: Where there is coral assign a value of 1 and NA otherwise**
```{r}
#### fasterize the polygons to our 3nm raster
fasterize_pal_py <- fasterize::fasterize(pal_py, raster = pal_3nm_rast , 
                                     field = NULL)

plot(fasterize_pal_py)

#### check to see if all points are NA and 1 for polygon raster
check_values_py <- getValues(fasterize_pal_py)
sum(check_values_py == 1, na.rm = TRUE)
# 4501
sum(is.na(check_values_py))
# 25392
unique(check_values_py)
# NA 1

```

**Calculate zonal stats with 3nm raster and fasterized Palmyra polygons. Convert to km^2**
```{r, echo = FALSE, eval = FALSE}
zonal_sums <- raster::zonal(fasterize_pal_py, 
                                     pal_3nm_rast,
                                     fun = "sum",
                                     na.rm = TRUE) #sum all coral cells for each ohi zone


zonal_sums_df <- data.frame(zonal_sums)

my_raster_res <- res(fasterize_pal_py)
my_raster_res ## 100 100

crs(fasterize_pal_py) ## units in m 
a <- area(fasterize_pal_py, na.rm = TRUE)
cell_size <- data.frame(area = a[!is.na(a)]) # each cell is 10000 m2? which means each cell is 0.01 km2  


zonal_sums_km2 <- zonal_sums_df %>%
  mutate(year = 2002, habitat = "coral",
         km2 = (0.01*sum)) %>% #one cell is equal to 0.01 km2
  dplyr::rename("rgn_id" = "zone") %>%
  dplyr::select(-sum)

## this shows 44.29 km2 compared to calculated GIS area of 145.2481 km2


```

```{r}
## try the other dataset found on NOAA
dir_PACIOOS <- file.path(file.path(dir_M, 'git-annex/foursite/palmyra/_raw_data/PACIOOS_data/pat_unepwcmc_all_coralreefs'))

unep_corals <- sf::st_read(dsn = file.path(dir_PACIOOS), layer = "pat_unepwcmc_all_coralreefs") %>%
  st_transform(crs = us_alb)

plot(unep_corals[2])

fasterize_unep_corals <- fasterize::fasterize(unep_corals, raster = pal_3nm_rast , 
                                     field = NULL)

plot(fasterize_unep_corals)


zonal_sums <- raster::zonal(fasterize_unep_corals, 
                                     pal_3nm_rast,
                                     fun = "sum",
                                     na.rm = TRUE) #sum all coral cells for each ohi zone


zonal_sums_df <- data.frame(zonal_sums)

my_raster_res <- res(fasterize_unep_corals)
my_raster_res ## 100 100

crs(fasterize_unep_corals) ## units in m 
a <- area(fasterize_unep_corals, na.rm = TRUE)
cell_size <- data.frame(area = a[!is.na(a)]) # each cell is 10000 m2? which means each cell is 0.01 km2  


zonal_sums_km2 <- zonal_sums_df %>%
  mutate(year = 2002, habitat = "coral",
         km2 = (0.01*sum)) %>% #one cell is equal to 0.01 km2
  dplyr::rename("rgn_id" = "zone") %>%
  dplyr::select(-sum)

## its the same data set... just a subset of the bigger one like we did above. 
```

